---
title: "Capelin_sdmTMB_cv"
output: html_document
date: "2026-07-06"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Problem: In SEBS in 2016 no capelin were caught at any stations which makes the spatiotemporal model for SEBS capelin impossible to converge. However, we can adjust the model form/input data to converge the model. 

Options: 
1. Use a time-varying random walk for year:

- a time-varying random walk for year (to use information from years before/after 2016 i.e., 2014 and 2018 to inform the year intercept for 2016). Otherwise it tries to esimate an unreasonable intercept for 2016. 

2. Use a reduced dataset without 2016

# libraries & functions
```{r}
# data pull function
source("Scripts/helper_fxns/clean_pull_data.R")

# make sure the proper packages are install and loaded
pkgs <- c("visreg", "sdmTMB", "sf", "concaveman", "cowplot", "gridExtra")

for (pkg in pkgs) {
  if (!require(pkg, character.only = TRUE)) {
    install.packages(pkg)
    library(pkg, character.only = TRUE)
  } else {
    message(paste("Package '", pkg, "' is already installed and loaded.", sep = ""))
  }
}
```

# bring in data
```{r}
tsns <- c(934083, 161979, 162035, 161980, 161976, 161977, 551209, 161975, 164711, 162041, 164708, 171672, 171671) # right now as i have this set up the tsns have to be these. Otherwise i need to adjust how they are combined in the clean_pull_data function
this.year <- 2025

ts <- pull_clean_catch(this.year = this.year, tsns = tsns, region = "SEBS") %>%
  # this pivot longer wont work if you change the tsns above 
  pivot_longer(cols = c("Chinook Salmon_J":"Forage"), names_to = "sp_name", values_to = "catch_kg") %>%
  add_utm_columns(., ll_names = c("Lon", "Lat"),
                  utm_names = c("X", "Y"), ll_crs = 4326, units = "km") %>%
  mutate(region = "SEBS", 
         crs = 32602)

# get the species names in a vector, there are 31, but we're only interested in a few for the ESRs 
species_names <- unique(ts$sp_name) 
species_names <- species_names[species_names %in% c("Capelin_All", "Forage", "Pacific Herring_All",
                                                   "Pollock_A0")]
ts <- filter(ts, sp_name %in% species_names) %>%
  filter(!(is.na(effort))) # drop any stations that have NA for effort

# datalist just by the species name so includes region for both within a specis
ts.list <- split(ts, list(ts$sp_name))
```

divide the data into just the sebs capelin dataframe
```{r}
ts.capelin <- filter(ts.list[["Capelin_All"]], region == "SEBS")
ts.capelin.red <- filter(ts.capelin, sample_year != 2016) # we know that the 2016 survey encountered no capelin so it explodes the estimates
```

# set up capelin sdmtmbs
```{r}
mesh.sebs.red <- make_mesh(ts.capelin.red, 
            xy_cols = c("X", "Y"),
            n_knots = 200)

mesh.sebs <- make_mes(ts.capelin, 
                      xy_cols = c("X", "Y"), 
                      n_knots = 200)

ts.capelin.red$log_effort <- log(ts.capelin.red$effort)
ts.capelin$log_effort <- log(ts.capelin$effort)
ts.capelin.red$fct_year <- as.factor(ts.capelin.red$sample_year)

# group data by years rather than grouping regardless of years
fold_vector <- as.numeric(as.factor(ts.capelin$sample_year))
fold_vector2 <- as.numeric(as.factor(ts.capelin.red$sample_year))
# this allows for leave-one-year-out validation
unique(ts.capelin$sample_year) # complete dataset with 2016
unique(ts.capelin.red$sample_year) # without 2016
```


# capelin cv

can we use cross validation to get a better idea of which modelling approach is better?
Following this vignette: https://sdmtmb.github.io/sdmTMB/articles/cross-validation.html
```{r}
TMB::openmp(n = 1)
# do cross validation four times, first with the full dataset then with the reduced dataset
# set up paralell processing
future::plan(future::multisession, workers = 4)
rw_cv <- sdmTMB_cv(data = ts.capelin, # trying out using the reduced dataset for both cross validation 
            formula = catch_kg ~ 0 + scale_doy,  # Remove the factor year
            offset = "log_effort",
            time_varying = ~ 1,                  # Turn on time-varying intercepts
            time_varying_type = "rw",
            time = "sample_year", # this needs to be an interger or numeric for random walk to function correctly. So it shouldn't be fct_year
            # remove extra time for cv only use in full model
            #extra_time =c(2013, 2015, 2017, 2019, 2020, 2021, 2023),
            mesh = mesh.sebs, # if red dataset chagne to mesh.sebs.red
            family = tweedie(link = "log"),
            spatial = "on",
            spatiotemporal = "iid",
            anisotropy = TRUE,
            fold_ids = fold_vector # if reduced change to fold_vector2
)

rw_cv_red <- sdmTMB_cv(data = ts.capelin.red, # trying out using the reduced dataset for both cross validation 
            formula = catch_kg ~ 0 + scale_doy,  # Remove the factor year
            offset = "log_effort",
            time_varying = ~ 1,                  # Turn on time-varying intercepts
            time_varying_type = "rw",
            time = "sample_year", # this needs to be an interger or numeric for random walk to function correctly. So it shouldn't be fct_year
            # remove extra time for cv only use in full model
            #extra_time =c(2013, 2015, 2017, 2019, 2020, 2021, 2023),
            mesh = mesh.sebs.red, # if red dataset chagne to mesh.sebs.red
            family = tweedie(link = "log"),
            spatial = "on",
            spatiotemporal = "iid",
            anisotropy = TRUE,
            fold_ids = fold_vector2 # if reduced change to fold_vector2
)


# 
# # Now set up your parallel structure
# this i think sets it up to run each model sequentially?
future::plan(future::sequential)
# run the reduced dataset model
rand_cv <- sdmTMB_cv(data = ts.capelin, 
            # right now estimating each yuear as a fixed effect
            #catch_kg ~ 0 + as.factor(sample_year) + scale_doy,
            # include a random effect for year
            catch_kg ~ 0 + scale_doy + (1 | fct_year),
            offset = "log_effort",
            time = "sample_year",
            mesh = mesh.sebs, 
            spatiotemporal = "iid",
            spatial = "on",
            family = tweedie(link = "log"),
            share_range = TRUE, # default for sdmTMB
            anisotropy = TRUE,
            fold_ids = fold_vector # change from fold vector to fold vector2 for red fdf
)

rand_cv_red <- sdmTMB_cv(data = ts.capelin.red, 
            # right now estimating each yuear as a fixed effect
            #catch_kg ~ 0 + as.factor(sample_year) + scale_doy,
            # include a random effect for year
            catch_kg ~ 0 + scale_doy + (1 | fct_year),
            offset = "log_effort",
            time = "sample_year",
            mesh = mesh.sebs.red, 
            spatiotemporal = "iid",
            spatial = "on",
            family = tweedie(link = "log"),
            share_range = TRUE, # default for sdmTMB
            anisotropy = TRUE,
            fold_ids = fold_vector2 # change from fold vector to fold vector2 for red fdf
)


rw_cv$sum_loglik # full dataset: -50205.22 # red df -53677.4
rw_cv_red$sum_loglik
rand_cv$sum_loglik # full dataset: -34874.86 # red df -35652.5, closer to 0 so better.
#red_cv$sum_loglik #(outputs -9.36e-21, -1.511e-20)
rand_cv_red$sum_loglik

# mean log likelihoods
mean_loglike_rw <- rw_cv$sum_loglik / nrow(rw_cv$data) # full df -51.177, reduced df -56.741
mean_loglike_rand <- red_cv$sum_loglik / nrow(red_cv$data) # -35.55032, reduced df -37.68

# RMSE across entire dataset:
sqrt(mean((rw_cv$data$catch_kg - rw_cv$data$cv_predicted)^2)) #full df 24.87; reduced df 27.929
sqrt(mean((rw_cv_red$data$catch_kg - rw_cv_red$data$cv_predicted)^2)) 
sqrt(mean((rand_cv$data$catch_kg - rand_cv$data$cv_predicted)^2)) #25.7013 # reduced df 26.021
sqrt(mean((rand_cv_red$data$catch_kg - rand_cv_red$data$cv_predicted)^2)) 

# ggplot() +
#   geom_point(aes(x = rw_cv$data$catch_kg, rw_cv$data$cv_predicted))

# RMSE by fold
rmse_cv_rw <- dplyr::group_by(rw_cv$data, cv_fold) |> 
  dplyr::summarize(
    mae = mean(abs(catch_kg - cv_predicted)),
    rmse = sqrt(mean((catch_kg - cv_predicted)^2)))
rmse_cv_rw_red <- dplyr::group_by(rw_cv_red$data, cv_fold) |> 
  dplyr::summarize(
    mae = mean(abs(catch_kg - cv_predicted)),
    rmse = sqrt(mean((catch_kg - cv_predicted)^2)))

# for random effects model
rmse_cv_rand <- dplyr::group_by(rand_cv$data, cv_fold) |> 
  dplyr::summarize(
    mae = mean(abs(catch_kg - cv_predicted)),
    rmse = sqrt(mean((catch_kg - cv_predicted)^2)))
rmse_cv_rand_red <- dplyr::group_by(rand_cv_red$data, cv_fold) |> 
  dplyr::summarize(
    mae = mean(abs(catch_kg - cv_predicted)),
    rmse = sqrt(mean((catch_kg - cv_predicted)^2)))

ggplot() +
  geom_line(aes(x = cv_fold, rmse), data = rmse_cv_rw, col = "red") +
  geom_line(aes(x = cv_fold, rmse), data = rmse_cv_rand, col = "blue")

# reduced dataset
ggplot() +
  geom_line(aes(x = cv_fold, rmse), data = rmse_cv_rw_red, col = "red") +
  geom_line(aes(x = cv_fold, rmse), data = rmse_cv_rand_red, col = "blue")
```

looks like overall the reduced random effects (which we'll treat as fixed effect in the final model) performs better than the random walk
so we'll refit both the random walk with all data and the reduced fixed effect model just to compare how the indicies perform. 

## set up final models
```{r}
# read in old mesh 
mesh.sebs <- make_mesh(ts.capelin,
                       xy_cols = c("X", "Y"),
                       n_knots = 200)
mesh.sebs.red <- make_mesh(ts.capelin.red, 
            xy_cols = c("X", "Y"),
            n_knots = 200)
```

```{r}
## fit model with random walk 
sebs.capelin.fit.rw <- sdmTMB(data = ts.capelin, # trying out using the reduced dataset for both cross validation 
            formula = catch_kg ~ 0 + scale_doy,  # Remove the factor year
            offset = ts.capelin$log_effort,
            time_varying = ~ 1,                  # Turn on time-varying intercepts
            time_varying_type = "rw",
            time = "sample_year", # this needs to be an interger or numeric for random walk to function correctly.
            extra_time =c(2013, 2015, 2017, 2019, 2020, 2021, 2023),
            mesh = mesh.sebs, 
            family = tweedie(link = "log"),
            spatial = "on",
            spatiotemporal = "iid",
            anisotropy = TRUE)


## fit model without 2016 
sebs.capelin.fit.red <- sdmTMB(data = ts.capelin.red, 
            # right now estimating each yuear as a fixed effect
            catch_kg ~ 0 + as.factor(sample_year) + scale_doy,
            offset = ts.capelin.red$log_effort,
            # if we wanted this to be random walk intercepts for year fct (time_varying = ~1)
            time = "sample_year",
            mesh = mesh.sebs.red, 
            spatiotemporal = "iid",
            spatial = "on",
            family = tweedie(link = "log"),
            share_range = TRUE, # default for sdmTMB
            # for the bering sea anisotropy important given the patterns across the shelf. 
            anisotropy = TRUE)

sanity(sebs.capelin.fit.red)
sanity(sebs.capelin.fit.rw)
tidy(sebs.capelin.fit.red, effects = "fixed") 
```
## prediction grids
read in the one created for the full dataset

```{r}
# read in full df prediction grid
sebs.grid <- read.csv("Results/2026_ESR/SEBS_prediction-grid.csv")

# create grid for reduced dataset
capelin.sf <- ts.capelin.red %>%
      dplyr::select(c(X, Y, eq_latitude, eq_longitude)) %>%
      mutate(X = X*1000, Y = Y*1000) %>% # convert to meters for the crs conversion
      st_as_sf(coords = c("X", "Y"), crs = unique(ts.capelin.red$crs))
    
convex_hull <- concaveman(capelin.sf, concavity = 3)
convex_buff <- st_buffer(convex_hull, dist = 15e3)
grid <- st_make_grid(convex_buff, cellsize = 10e3, what = "centers") %>%
  st_intersection(convex_buff)

# create spatial grid
bs.grid <- do.call(rbind, st_geometry(grid)) %>%
  as.tibble() %>% setNames(c("X", "Y")) %>%
  # convert df m back to km
  mutate(X = X/1e3, Y = Y/1e3)
# replicate the grid across years (always in model)
bs.grid.yrs <- replicate_df(bs.grid, "sample_year", unique(ts.capelin.red$sample_year))
# assuming we have scale_doy in model (which we should) replicate with the scaled doy
# to predict on the median value of the survey
bs.grid.doy <- replicate_df(bs.grid.yrs, "scale_doy", 0)
```

# predict indicies
```{r}
pred.capelin.rw <- predict(sebs.capelin.fit.rw, newdata = sebs.grid, return_tmb_object = T)
pred.capelin.red <- predict(sebs.capelin.fit.red, newdata = bs.grid.doy, return_tmb_object = T)

capelin.index.rw <- get_index(pred.capelin2, bias_correct = TRUE)
capelin.index.red <- get_index(pred.capelin, bias_correct = TRUE)
```


compare the two capelin indicies
```{r}
capelin.index.rw$model_type <- "random_walk"
capelin.index.red %>%
  mutate(model_type = "reduced_model") %>%
  rbind(capelin.index.rw) %>%
  ggplot() +
  geom_pointrange(aes(x = sample_year, y = est, 
                      ymin = lwr, ymax = upr,
                      color = model_type))

# in general the random walk model provides an estimate for 2016 (which is 0) but pushes the estimes around the peak capelin years up and reduces the variability given the extra information that contributing to the estimate (years before/after).
#' the reduced model has larger ci for the years where capelin peaked because of how intermittant they were encounterd. High encounter rate or none at all. boom/bust situation. 
#' Overall based on the cross validation it is slightly better to use the reduced model over the random wlak. 
```

