---
title: "fit_sdm_models"
output: html_document
date: "2026-06-25"
editor_options: 
  chunk_output_type: console
---

Pull in fxns
```{r}
source("Scripts/helper_fxns/clean_pull_data.R")
```

# libraries
```{r}
# make sure the proper packages are install and loaded
pkgs <- c("visreg", "sdmTMB", "sf", "concaveman", "cowplot", "gridExtra")

for (pkg in pkgs) {
  if (!require(pkg, character.only = TRUE)) {
    install.packages(pkg)
    library(pkg, character.only = TRUE)
  } else {
    message(paste("Package '", pkg, "' is already installed and loaded.", sep = ""))
  }
}
```

# Pull in data
```{r}
tsns <- c(934083, 161979, 162035, 161980, 161976, 161977, 551209, 161975, 164711, 162041, 164708, 171672, 171671) # right now as i have this set up the tsns have to be these. Otherwise i need to adjust how they are combined in the clean_pull_data function
this.year <- 2025

nbs.ts <- pull_clean_catch(this.year = this.year, tsns = tsns, region = "NBS") %>%
  # this pivot longer wont work if you change the tsns above 
  pivot_longer(cols = c("Chinook Salmon_J":"Forage"), names_to = "sp_name", values_to = "catch_kg") %>%
  add_utm_columns(., ll_names = c("Lon", "Lat"),
                  utm_names = c("X", "Y"), ll_crs = 4326, units = "km") %>%
  mutate(region = "NBS",
         crs = 32603)

sebs.ts <- pull_clean_catch(this.year = this.year, tsns = tsns, region = "SEBS") %>%
  # this pivot longer wont work if you change the tsns above 
  pivot_longer(cols = c("Chinook Salmon_J":"Forage"), names_to = "sp_name", values_to = "catch_kg") %>%
  add_utm_columns(., ll_names = c("Lon", "Lat"),
                  utm_names = c("X", "Y"), ll_crs = 4326, units = "km") %>%
  mutate(region = "SEBS", 
         crs = 32602)

# glimpse(nbs.ts)
# glimpse(sebs.ts)
# combine and then split
ts <- rbind(sebs.ts, nbs.ts) # crs for NBS is 32603, SEBS is 32602
  
# get the species names in a vector, there are 31, but we're only interested in a few for the ESRs 
species_names <- unique(ts$sp_name) 
species_names <- species_names[species_names %in% c("Capelin_All", "Forage", "Pacific Herring_All",
                                                   "Pollock_A0")]
ts <- filter(ts, sp_name %in% species_names) %>%
  filter(!(is.na(effort))) # drop any stations that have NA for effort

# datalist just by the species name so includes region for both within a specis
ts.list <- split(ts, list(ts$sp_name))

```

## understand the catch / encounter data for each region and species
Things to look for -
- 0% or 100% encounter rate for years 

```{r}
# check to make sure there's no negative or NAs in catch data
table(is.na(ts$catch_kg)) # none
table(ts$catch_kg < 0) # none
table(is.na(ts$effort)) # any NAs in effort need to be removed

forage <- ts.list[["Forage"]]
pol <- ts.list[["Pollock_A0"]]
capelin <- ts.list[["Capelin_All"]]
herring <- ts.list[["Pacific Herring_All"]]

# both regions here
# some years with zero, 0 encounter rates, i.e., multiple years where all stations had some forage catch
# foro NBS very few zero encounters, SEBS more 0 encounters but very few
table(forage$catch_kg > 0, forage$sample_year, forage$region)
# for NBS good mix of pol encounter/non encounter, for SEBS years where no surveys done but otherwise a good mix in early years for encounter non encounter, but later years only positive encounter mainly (reduced survey footprint meaning only middle domain stations with pollock)
table(pol$catch_kg > 0, pol$sample_year, pol$region)
# for NBS good mix of capelin encounter/non encounter in early years, with later years a lot fewer capelin encountered (especially 2017-2021). For SEBS, even fewer encounters across the board especially in recent years (2016 no encounters etc)
table(capelin$catch_kg > 0, capelin$sample_year, capelin$region)
# for NBS good mix of encounter non encounter for both NBS and SEBS
table(herring$catch_kg >0, herring$sample_year, herring$region)


#' conclusions:
#' Forage: perhaps no need for a hurdle model, just use a positive encounter model for catch such as the tweedie
#' Pollock: NBS likely needs a hurdle model, like delta_gamma or delta_lognormal, SEBS can be tweedie
#' Capelin, high zero encounters probably delta_lognormal
#' Herring probably can use a delta-gamma or delta_lognormal. 

```

testing variables
```{r}
# sp_name <- "Forage"
# survey_region <- "NBS"
# run_fast <- TRUE
# n_x <- 200
# family = tweedie(link = "log")
# #family = delta_gamma(link1 = "logit", link2 = "log")
# spatial = "off"
# spatiotemporal = "iid"
# #spatial = list("off", "on") # on, on is equivalent (with spatiotemporal arg) to Fieldconfig 1,1,1,1
# #spatiotemporal = list("off", "iid")
# data = ts.list
```


# sdmTMB
```{r}
# this function fits a sdmTMB model that has offset log effort and a model structure of f-year and s-scale.doy
# this has indepedent year effects
set_fit_TMB <- function(
    data,
    survey_region,
    sp_name,
    run_fast = FALSE,
    n_x = 200,
    spatiotemporal = "IID",
    spatial = "on",
    family = tweedie(link = "log")
) {
  
  set.seed(2026)
  
  # set directory for results
  today <- Sys.Date()
  year <- year(today)
  path <- here("Results", paste0(year, "_ESR"), paste0(sp_name, "_", survey_region), today)
  if(!dir.exists(path)) dir.create(path, recursive = T)
  
  # change knots and if running quickly
  # (i.e. not a final model)
   if(run_fast == TRUE){
    n_x <- 50
  } else {
    # if we want to have different models
    if(survey_region == "NBS") {
      n_x = 200
    } else {
      if(survey_region == "SEBS") {
        n_x = 200
      }
    }
  }
  
  
  # change data pending region & sp name
  if(survey_region == "NBS"){
    ts.tmp <- ts.list[[sp_name]]
    ts <- filter(ts.tmp, region == survey_region) %>%
      as.data.frame()
  } else {
    if(survey_region == "SEBS") {
      ts.tmp <- ts.list[[sp_name]]
    ts <- filter(ts.tmp, region == survey_region) %>%
      as.data.frame()
    } else {stop("data not supplied for region")}
    
  }
  
  # print out median doy for *knowledge*
  ts.doy <- ts %>%
    group_by(station_id) %>%
    mutate(doy = yday(haul_date)) %>%
    summarise(mean.doy = mean(doy)) %>%
    summarise(median.doy = median(mean.doy),
              mean = mean(mean.doy))
  
  print(paste(survey_region, sp_name, "median doy", ts.doy$median.doy))
  
  ## create mesh
  mesh <- make_mesh(ts, 
            xy_cols = c("X", "Y"),
            n_knots = n_x)
  
  ## set up the effort offset 
  ts$log_effort <- log(ts$effort)

  ## fit model 
  fit <- sdmTMB(data = ts, 
                # right now estimating each yuear as a fixed effect
                catch_kg ~ 0 + as.factor(sample_year) + scale_doy,
                offset = ts$log_effort,
                # if we wanted this to be random walk intercepts for year fct (time_varying = ~1)
                time = "sample_year",
                mesh = mesh, 
                spatiotemporal = spatiotemporal,
                spatial = spatial,
                family = family,
                share_range = TRUE, # default for sdmTMB
                # for the bering sea anisotropy important given the patterns across the shelf. 
                anisotropy = TRUE)

  print(sanity(fit))
  
  # pulled from lauren's code
  # is the model fit?
  if(inherits(fit, "try-error")) {
    print(paste("model error for species", sp_name, "in region", survey_region))
    
  } else {
    # otherwise, check to make sure the sanity checks are met
    if(sanity(fit)$all_ok == FALSE) fit <- try(run_extra_optimization(fit)) }
  # after that, check to see if the
  if(inherits(fit, "try-error")) {
    print(paste("model error for species", sp_name, "in region", survey_region)) }
  else {
    # if the model doesn't have any sanity issues or model issues then... 
    
    # save model object
    saveRDS(fit, file = paste0(here(path, sp_name), "_fit.RDS"))
    
    # look to see if there is a prediction grid already in this years ESR folder
    if(file.exists(here("Results", paste0(year, "_ESR"),
                                       paste0(survey_region, "_prediction-grid.csv")))) {
      print(paste("prediction grid for", survey_region, "already exists"))
    } else {
      # create prediction grids for this model
    # take the data input, convert to sf 
    ts.sf <- ts %>%
      dplyr::select(c(X, Y, eq_latitude, eq_longitude)) %>%
      mutate(X = X*1000, Y = Y*1000) %>% # convert to meters for the crs conversion
      st_as_sf(coords = c("X", "Y"), crs = unique(ts$crs))
    
    convex_hull <- concaveman(ts.sf, concavity = 3)
    convex_buff <- st_buffer(convex_hull, dist = 15e3)
    grid <- st_make_grid(convex_buff, cellsize = 10e3, what = "centers") %>%
      st_intersection(convex_buff)
    
    # create spatial grid
    bs.grid <- do.call(rbind, st_geometry(grid)) %>%
      as.tibble() %>% setNames(c("X", "Y")) %>%
      # convert df m back to km
      mutate(X = X/1e3, Y = Y/1e3)
    # replicate the grid across years (always in model)
    bs.grid.yrs <- replicate_df(bs.grid, "sample_year", unique(ts$sample_year))
    # assuming we have scale_doy in model (which we should) replicate with the scaled doy
    # to predict on the median value of the survey
    bs.grid.doy <- replicate_df(bs.grid.yrs, "scale_doy", 0)
    
    write.csv(bs.grid.doy, file = here("Results", paste0(year, "_ESR"),
                                       paste0(survey_region, "_prediction-grid.csv")),
              row.names = FALSE)
    } # close prediction grid if else creation
    
    
    
  } # close checking model errors
  
  print(paste("Model object and prediction grid (by year and scaled_doy and space) saved to", paste0(here(path))))
  
  return(fit)
  
}


```


```{r eval=FALSE, include=FALSE}
# ObsModel 2,3 -> family = delta_gamma(link1 = "logit", link2 = "log")?
# FieldConfig, first two elements are the spatial/spatiotemporal structure for the presence / absence part of the model. The second two elements are the spatial/spatiotemporal structure for the encounter model. 

# other considerations, share_range in sdmTMB
# scale / range parameters are independently estimated for each random field but sdmTMB shares it (with share_range = T)

nbs.forage.fit <- set_fit_TMB(data = ts.list, 
                           survey_region = "NBS", 
                           sp_name = "Forage", 
                           run_fast = FALSE, 
                           family = tweedie(link = "log"),
                           spatial = "on",
                           spatiotemporal = "iid",
                           n_x = 200)

nbs.pollock.fit <- set_fit_TMB(data = ts.list, 
                           survey_region = "NBS", 
                           sp_name = "Pollock_A0", 
                           run_fast = FALSE, 
                           family = tweedie(link = "log"),
                           spatial = "on",
                           spatiotemporal = "iid",
                           n_x = 200)

nbs.herring.fit <- set_fit_TMB(data = ts.list,
                                survey_region = "NBS",
                                sp_name = "Pacific Herring_All",
                                run_fast = FALSE,
                                family = delta_gamma(link1 = "logit", link2 = "log"),
                                # spatial = list(on, on) is equivalent to FieldConfig (1, X, 1, X)
                                spatial = list("on", "on"),
                                # list(iid, iid) equivalent to FieldConfig (X, 1, X, 1)
                                spatiotemporal = list("off", "iid"), 
                                n_x = 200) # fieldconfig (1,0,1,1)

nbs.capelin.fit <- set_fit_TMB(data = ts.list,
                                survey_region = "NBS",
                                sp_name = "Capelin_All",
                                run_fast = FALSE,
                                family = delta_gamma(link1 = "logit", link2 = "log"),
                                # spatial = list(on, on) is equivalent to FieldConfig (1, X, 1, X)
                                spatial = list("on", "on"),
                                # list(iid, iid) equivalent to FieldConfig (X, 1, X, 1)
                                spatiotemporal = list("iid", "iid"), 
                                n_x = 200) # fieldconfig (1,0,1,1)

sebs.forage.fit <- set_fit_TMB(data = ts.list, 
                           survey_region = "SEBS", 
                           sp_name = "Forage", 
                           run_fast = FALSE, 
                           family = tweedie(link = "log"),
                           spatial = "on",
                           spatiotemporal = "iid",
                           n_x = 200)

sebs.pollock.fit <- set_fit_TMB(data = ts.list, 
                           survey_region = "SEBS", 
                           sp_name = "Pollock_A0", 
                           run_fast = FALSE, 
                           family = tweedie(link = "log"),
                           spatial = "on",
                           spatiotemporal = "iid",
                           n_x = 200)

# takes awhile to run
sebs.herring.fit <- set_fit_TMB(data = ts.list,
                                survey_region = "SEBS",
                                sp_name = "Pacific Herring_All",
                                run_fast = FALSE,
                                # equivalent to ObsModel = c(2,3)
                                family = delta_gamma(link1 = "logit", link2 = "log"),
                                # spatial = list(on, on) is equivalent to FieldConfig (1, X, 1, X)
                                spatial = list("on", "off"),
                                # equivalent to FieldConfig (X, 1, X, 1)
                                spatiotemporal = list("iid", "iid"), 
                                n_x = 200) # with spatial on, off and iid iid its = 1,1,0,1

# this  model doesn't converge it requires more sophisticated adjustments to the model form
# sebs.capelin.fit <- set_fit_TMB(data = map2(ts.list, names(ts.list), ~ filter(.x, sample_year != 2016)),
#                                 # above we remove 2016 from the list (all sp)
#                                 survey_region = "SEBS",
#                                 sp_name = "Capelin_All",
#                                 run_fast = FALSE,
#                                 family = tweedie(link = "log"),
#                                 # spatial = list(on, on) is equivalent to FieldConfig (1, X, 1, X)
#                                 spatial = "on",
#                                 # list(iid, iid) equivalent to FieldConfig (X, 1, X, 1)
#                                 spatiotemporal = "off",
#                                 n_x = 200) #


```

specifically for capelin in sebs we're looking at the extreme outliers or zero inflation
```{r}
# instead of using my function that has consistent model structure we're going to refit it here with a random walk for years and compare it with removing the year all together

ts.capelin <- filter(ts.list[["Capelin_All"]], region == "SEBS")
ts.capelin.red <- filter(ts.capelin, sample_year != 2016) # we know that the 2016 survey encountered no capelin so it explodes the estimates

mesh.sebs.red <- make_mesh(ts.capelin.red, 
            xy_cols = c("X", "Y"),
            n_knots = 200)

mesh.sebs <- make_mesh(ts.capelin, 
            xy_cols = c("X", "Y"),
            n_knots = 200)

ts.capelin.red$log_effort <- log(ts.capelin.red$effort)
ts.capelin$log_effort <- log(ts.capelin$effort)
ts.capelin$fct_year <- as.factor(ts.capelin$sample_year)

## fit model without 2016 
sebs.capelin.fit.red <- sdmTMB(data = ts.capelin.red, 
            # right now estimating each yuear as a fixed effect
            catch_kg ~ 0 + as.factor(sample_year) + scale_doy,
            offset = ts.capelin.red$log_effort,
            # if we wanted this to be random walk intercepts for year fct (time_varying = ~1)
            time = "sample_year",
            mesh = mesh.sebs.red, 
            spatiotemporal = "iid",
            spatial = "on",
            family = tweedie(link = "log"),
            share_range = TRUE, # default for sdmTMB
            # for the bering sea anisotropy important given the patterns across the shelf. 
            anisotropy = TRUE)

sanity(sebs.capelin.fit.red)
tidy(sebs.capelin.fit.red, effects = "fixed") # estimates and standard errors look a lot better

# fit model with a time varying random walk
sebs.capelin.fit2 <- sdmTMB(
            data = ts.capelin,
            formula = catch_kg ~ 0 + scale_doy,  # Remove the factor year
            offset = ts.capelin$log_effort,
            time_varying = ~ 1,                  # Turn on time-varying intercepts
            time_varying_type = "rw",
            time = "sample_year",
            extra_time =c(2013, 2015, 2017, 2019, 2020, 2021, 2023),
            mesh = mesh.sebs,
            family = tweedie(link = "log"),
            spatial = "on",
            spatiotemporal = "off",
            anisotropy = TRUE)

sanity(sebs.capelin.fit2)
tidy(sebs.capelin.fit2)

# AIC doesn't make sense because there are different underpinning data for the models
AIC(sebs.capelin.fit.red)
AIC(sebs.capelin.fit2)
```

have to recreate the prediction grid 
```{r}
capelin.sf <- ts.capelin.red %>%
      dplyr::select(c(X, Y, eq_latitude, eq_longitude)) %>%
      mutate(X = X*1000, Y = Y*1000) %>% # convert to meters for the crs conversion
      st_as_sf(coords = c("X", "Y"), crs = unique(ts.capelin.red$crs))
    
    convex_hull <- concaveman(capelin.sf, concavity = 3)
    convex_buff <- st_buffer(convex_hull, dist = 15e3)
    grid <- st_make_grid(convex_buff, cellsize = 10e3, what = "centers") %>%
      st_intersection(convex_buff)
    
    # create spatial grid
    bs.grid <- do.call(rbind, st_geometry(grid)) %>%
      as.tibble() %>% setNames(c("X", "Y")) %>%
      # convert df m back to km
      mutate(X = X/1e3, Y = Y/1e3)
    # replicate the grid across years (always in model)
    bs.grid.yrs <- replicate_df(bs.grid, "sample_year", unique(ts.capelin.red$sample_year))
    # assuming we have scale_doy in model (which we should) replicate with the scaled doy
    # to predict on the median value of the survey
    bs.grid.doy <- replicate_df(bs.grid.yrs, "scale_doy", 0)
```



# Predict biomass across prediction grid and calculate index biomass
Read in correct grid/fitted models
```{r eval=FALSE, include=FALSE}
# read in prediction grid for said survey region 
nbs.grid <- read.csv(here("Results", 
                          paste0(year(Sys.Date()), "_ESR"), "NBS_prediction-grid.csv"))
sebs.grid <- read.csv(here("Results", 
                           paste0(year(Sys.Date()), "_ESR"), "SEBS_prediction-grid.csv"))

# predict sebs and nbs
# model list 
model.list <- list(nbs.forage.fit, nbs.pollock.fit, nbs.herring.fit, nbs.capelin.fit,
                   sebs.forage.fit, sebs.pollock.fit, sebs.herring.fit)
names(model.list) <- c("nbs.forage", "nbs.pollock", "nbs.herring", "nbs.capelin",
                       "sebs.forage", "sebs.pollock", "sebs.herring")

grid.list <- list(nbs.grid, sebs.grid)
names(grid.list) <- c("nbs", "sebs")

#predict & extract indicies
predict_index <- function(model.name, model.list, grid.list) {
  mod <- model.list[[model.name]]
  
  #is there a nbs at beginning of the model name?
  if(str_detect(model.name, "^nbs\\.")) {
    # make the grid the nbs grid
    grid <- grid.list[["nbs"]]
  } else {
    grid <- grid.list[["sebs"]]
    
  }
  
  preds <- predict(mod, newdata = grid, return_tmb_object = TRUE)
  index <- get_index(preds, bias_correct = TRUE)
  
  # column with model name
  index$model <- model.name
  print(model.name)
  return(index)
  
}

# if you want to use parallel processing with increased cores
TMB::openmp(n = 8) # can use parallel::detectCores to see how many cores the computer has

index_results <- map(names(model.list), 
    ~ predict_index(model.name = .x, model.list = model.list, grid.list = grid.list))

indicies <- bind_rows(index_results) %>%
  mutate(region = ifelse(str_detect(model, "^nbs\\."), "NBS", "SEBS"),
         species = ifelse(str_detect(model, "forage"), "Forage", 
                          ifelse(str_detect(model, "pollock"), "Pollock",
                                           ifelse(str_detect(model, "capelin"), "Capelin",
                                                             ifelse(str_detect(model, "herring"), "Herring", NA)))))

# capelin model for sebs were fit differently than the other models
pred.capelin2 <- predict(sebs.capelin.fit2, newdata = sebs.grid, return_tmb_object = T)
pred.capelin <- predict(sebs.capelin.fit.red, newdata = bs.grid.doy, return_tmb_object = T)

capelin.index2 <- get_index(pred.capelin2, bias_correct = TRUE)
capelin.index <- get_index(pred.capelin, bias_correct = TRUE)

# once we determine which capelin model is correct add it into the indicies df
write.csv(indicies, here("Results", 
                         paste0(year(Sys.Date()), "_ESR"), 
                         paste0("Index_rel_biomass_", Sys.Date(), ".csv")))

```

# plot the ESR indicies

```{r}
# if they need to be loaded can be done here
# include the correct date of the version of relative abundance  you want to use
correct.date <- "2026-07-01" # most recent run
indicies <- read.csv(here("Results", paste0(year(Sys.Date()), "_ESR"),
              paste0("Index_rel_biomass_", correct.date, ".csv")))

sp.list <- unique(indicies$species)

plot_ESR <- function(sp_name, data) {
  dat <- filter(data, species == sp_name)
  
  ggplot(aes(x = sample_year, y = est, group = region, 
                          color = region), data = dat) +
          geom_point(size=2)+
          geom_line()+
          geom_errorbar(aes(ymin = lwr, ymax = upr,
                            group = region, color = region), width=.1, 
                        position=position_dodge(0.05)) +
          theme_cowplot(12)+
          ylab(bquote("CPUE (kg/"* ~ km^2*')'))+
          labs(title=paste(sp_name, "- Late Summer/Fall Surface Trawl Surveys"), 
               x ="Year", color = "Region", cex=3)+
          theme(legend.position = "right",
                axis.title = element_text(size = 16),
                legend.text = element_text(size = 12),
                legend.title = element_text(size = 16),
                axis.text = element_text(size = 12),
                panel.background = element_rect(colour = "black", linewidth=1))+
          #scale_y_continuous(limits = c(0, 650)) +
          scale_color_manual(values=c('#003366','#0099CC'))+
          #scale_color_manual(values=c('#003366','#0099CC'))+
          geom_point(size=1)
  
}


plot_list <- list()
for(i in 1:length(sp.list)) {
  sp_name <- sp.list[[i]]
  plot_list[[i]] <- plot_ESR(sp_name = sp_name, data = indicies)
}

gridExtra::grid.arrange(grobs = plot_list)


```


compare the two capelin indicies
```{r}
capelin.index2$model_type <- "random_walk"
capelin.index %>%
  mutate(model_type = "reduced_model") %>%
  rbind(capelin.index2) %>%
  ggplot() +
  geom_pointrange(aes(x = sample_year, y = est, 
                      ymin = lwr, ymax = upr,
                      color = model_type))
```

# capelin cv

can we use cross validation to get a better idea of which modelling approach is better?
Following this vignette: https://sdmtmb.github.io/sdmTMB/articles/cross-validation.html
```{r}
# group data by years rather than slicing years
fold_vector <- as.numeric(as.factor(ts.capelin$sample_year))
# fold_vector2 <- as.numeric(as.factor(ts.capelin.red$sample_year))
# this allows for leave-one-year-out validation

# set up paralell processing
future::plan(future::multisession)
rw_cv <- sdmTMB_cv(data = ts.capelin,
            formula = catch_kg ~ 0 + scale_doy,  # Remove the factor year
            offset = "log_effort",
            time_varying = ~ 1,                  # Turn on time-varying intercepts
            time_varying_type = "rw",
            time = "sample_year",
            # remove extra time for cv 
            #extra_time =c(2013, 2015, 2017, 2019, 2020, 2021, 2023),
            mesh = mesh.sebs,
            family = tweedie(link = "log"),
            spatial = "on",
            spatiotemporal = "off",
            anisotropy = TRUE,
            fold_ids = fold_vector
)

TMB::openmp(n = 1) 

# Now set up your parallel structure
future::plan(future::sequential)
# run the reduced dataset model
red_cv <- sdmTMB_cv(data = ts.capelin, 
            # right now estimating each yuear as a fixed effect
            #catch_kg ~ 0 + as.factor(sample_year) + scale_doy,
            # include a random effect for year
            catch_kg ~ 0 + scale_doy + (1 | fct_year),
            offset = "log_effort",
            time = "fct_year",
            mesh = mesh.sebs, 
            spatiotemporal = "iid",
            spatial = "on",
            family = tweedie(link = "log"),
            share_range = TRUE, # default for sdmTMB
            anisotropy = TRUE,
            fold_ids = fold_vector
)



rw_cv$sum_loglik # full dataset: -33453.78
red_cv$sum_loglik # full dataset: -34874.86
#red_cv$sum_loglik #(outputs -9.36e-21, -1.511e-20)

# mean log likelihoods
mean_loglike_rw <- rw_cv$sum_loglik / nrow(rw_cv$data) # full df -34.10172
mean_loglike_red <- red_cv$sum_loglik / nrow(red_cv$data) # -35.55032

# RMSE across entire dataset:
sqrt(mean((rw_cv$data$catch_kg - rw_cv$data$cv_predicted)^2)) #24.92
sqrt(mean((red_cv$data$catch_kg - red_cv$data$cv_predicted)^2)) #0

# ggplot() +
#   geom_point(aes(x = rw_cv$data$catch_kg, rw_cv$data$cv_predicted))

# RMSE by fold
rmse_cv <- dplyr::group_by(rw_cv$data, cv_fold) |> 
  dplyr::summarize(
    mae = mean(abs(catch_kg - cv_predicted)),
    rmse = sqrt(mean((catch_kg - cv_predicted)^2)))
# for random effects model
rmse_cv_red <- dplyr::group_by(red_cv$data, cv_fold) |> 
  dplyr::summarize(
    mae = mean(abs(catch_kg - cv_predicted)),
    rmse = sqrt(mean((catch_kg - cv_predicted)^2)))

ggplot() +
  #geom_line(aes(x = cv_fold, rmse), data = rmse_cv, col = "red") +
  geom_line(aes(x = cv_fold, rmse), data = rmse_cv_red, col = "blue")
```

