---
title: "fit_sdm_models"
output: html_document
date: "2026-06-25"
editor_options: 
  chunk_output_type: console
---

Pull in fxns
```{r}
source("Scripts/helper_fxns/clean_pull_data.R")
```

# libraries
```{r}
# make sure the proper packages are install and loaded
pkgs <- c("visreg", "sdmTMB", "sf", "concaveman", "cowplot", "gridExtra", "patchwork",
          "lubridate", "here", "ggplotify", "tidyverse")

for (pkg in pkgs) {
  if (!require(pkg, character.only = TRUE)) {
    install.packages(pkg)
    library(pkg, character.only = TRUE)
  } else {
    message(paste("Package '", pkg, "' is already installed and loaded.", sep = ""))
  }
}
```

# Pull in data
```{r}
tsns <- c(934083, 161979, 162035, 161980, 161976, 161977, 551209, 
          161975, 164711, 162041, 164708, 171672, 171671,
            # add in the jellyfish tsns (below)
          50623, 51640, 51641, 51669, 51671, 51695, 51696, 51700, 51701, 51705, 51707, 719327) # right now as i have this set up the tsns have to be these. Otherwise i need to adjust how they are combined in the clean_pull_data function which is dumb but I dont have the energy to fix

# adjust to the current year
this.year <- 2025

nbs.ts <- pull_clean_catch(this.year = this.year, tsns = tsns, region = "NBS") %>%
  # this pivot longer wont work if you change the tsns above 
  pivot_longer(cols = c("Chinook Salmon_J":"Jellyfish"), names_to = "sp_name", values_to = "catch_kg") %>%
  add_utm_columns(., ll_names = c("Lon", "Lat"),
                  utm_names = c("X", "Y"), ll_crs = 4326, units = "km") %>%
  mutate(region = "NBS",
         crs = 32603)

sebs.ts <- pull_clean_catch(this.year = this.year, tsns = tsns, region = "SEBS") %>%
  # this pivot longer wont work if you change the tsns above 
  pivot_longer(cols = c("Chinook Salmon_J":"Jellyfish"), names_to = "sp_name", values_to = "catch_kg") %>%
  add_utm_columns(., ll_names = c("Lon", "Lat"),
                  utm_names = c("X", "Y"), ll_crs = 4326, units = "km") %>%
  mutate(region = "SEBS", 
         crs = 32602)

# glimpse(nbs.ts)
# glimpse(sebs.ts)
# combine and then split
ts <- rbind(sebs.ts, nbs.ts) # crs for NBS is 32603, SEBS is 32602
  
# get the species names in a vector, there are 31, but we're only interested in a few for the ESRs 
species_names <- unique(ts$sp_name) 
species_names <- species_names[species_names %in% c("Capelin_All", "Forage", "Pacific Herring_All",
                                                   "Pollock_A0", "Pacific Cod_A0", "Jellyfish")]
ts <- filter(ts, sp_name %in% species_names) %>%
  filter(!(is.na(effort))) # drop any stations that have NA for effort, right now drops 5

ts <- mutate(ts, log_effort = log(effort)) %>%
  group_by(region, sp_name) %>%
  mutate(median_doy = median(doy),
         scale_doy = doy - median_doy,
         mean_doy = mean(doy),
         sd_doy = sd(doy),
         scale_doy2 = (doy - mean_doy)/sd_doy,
         scaled_median = (median_doy - mean_doy) / sd_doy)

# remove the 2016 year for capelin and 2003 for jellyfish (no jellyfish recorded in 2003)
# remove an instance where the pcod caught in 2012 at a station wasn't weighed and so uses a placeholder of -999
# also tried to remove 0 catch years for pcod in NBS to get model to converge but honestly that 
# model sucks and shouldn't be used. 
ts.red <- filter(ts, !(region == "SEBS" & sp_name == "Capelin_All" & sample_year == 2016),
                     !(region %in% c("NBS", "SEBS") & sp_name == "Jellyfish" & sample_year == 2003),
                     !(region == "NBS" & sp_name == "Pacific Cod_A0" & 
                         sample_year %in% c(2003, 2008,2009, 2010,2014, 2018)),
                     catch_kg >= 0)
#write.csv(ts.red, here("Results", paste0(year(Sys.Date()), "_ESR"), paste0("nbs_sebs_timeseries.csv")))

# datalist just by the species name so includes region for both within a specis
ts.list <- split(ts.red, list(ts.red$sp_name))

```

## understand the catch / encounter data for each region and species
Things to look for -
- 0% or 100% encounter rate for years 

```{r}
# check to make sure there's no negative or NAs in catch data
table(is.na(ts.red$catch_kg)) # none
table(ts.red$catch_kg < 0) # none
table(is.na(ts.red$effort)) # any NAs in effort need to be removed

forage <- ts.list[["Forage"]]
pol <- ts.list[["Pollock_A0"]]
capelin <- ts.list[["Capelin_All"]]
herring <- ts.list[["Pacific Herring_All"]]
pcod <- ts.list[["Pacific Cod_A0"]]
jelly <- ts.list[["Jellyfish"]]

# both regions here
# some years with zero, 0 encounter rates, i.e., multiple years where all stations had some forage catch
# foro NBS very few zero encounters, SEBS more 0 encounters but very few
table(forage$catch_kg > 0, forage$sample_year, forage$region)
# for NBS good mix of pol encounter/non encounter, for SEBS years where no surveys done but otherwise a good mix in early years for encounter non encounter, but later years only positive encounter mainly (reduced survey footprint meaning only middle domain stations with pollock)
table(pol$catch_kg > 0, pol$sample_year, pol$region)
# for NBS good mix of capelin encounter/non encounter in early years, with later years a lot fewer capelin encountered (especially 2017-2021). For SEBS, even fewer encounters across the board especially in recent years (2016 no encounters etc)
table(capelin$catch_kg > 0, capelin$sample_year, capelin$region)
# for NBS good mix of encounter non encounter for both NBS and SEBS
table(herring$catch_kg >0, herring$sample_year, herring$region)
# for pcod lots of instances of no encounter in the nbs very few years with positive enounters
# this will cause conversion issues which will likely need to be solved with a random walk and/or dropping data. For sebs more years with positive encountersno years without some enounters so thats good
table(pcod$catch_kg > 0, pcod$sample_year, pcod$region)
# nearly no stations that didn't have some biomass of jellyfish (although some which feels improbable)
table(jelly$catch_kg > 0, jelly$sample_year, jelly$region)

#' conclusions:
#' Forage: perhaps no need for a hurdle model, just use a positive encounter model for catch such as the tweedie
#' Pollock: NBS likely needs a hurdle model, like delta_gamma or delta_lognormal, SEBS can be tweedie
#' Capelin, high zero encounters probably delta_lognormal
#' Herring probably can use a delta-gamma or delta_lognormal. 
#' Pcod NBS will need a hurdle model probably with a random walk for years, SEBS try out a tweedie but then can come back to a hurdle if needed
#' Jellyfish: tweedie

print(pol %>% filter(region == "NBS",
               catch_kg > 0) %>%
  count(sample_year), n = 21)
```


# sdmTMB wrapper fxn
```{r}
# function that creates a prediction grid for each species
create_prediction_grid <- function(ts, fit, survey_region, sp_name, year, ts_scaled_median) {
  
  # While not necessary to have each species with its own prediction grid we have several species
  # that have different length time series so having indidivudal preidction grids are good. 
  # This ensures custom year subsets never overwrite or share files with other species.
  grid_file <- here(
    "Results", 
    paste0(year, "_ESR"), 
    "Prediction_grids",
    paste0(survey_region, "_", sp_name, "_prediction-grid.csv")
  )
  
  grid_folder <- here(
    "Results", 
    paste0(year, "_ESR"), 
    "Prediction_grids"
    )
  
  if(!dir.exists(grid_folder)) dir.create(grid_folder, recursive = T)
  
  # If THIS specific species' grid already exists, skip it
  if (file.exists(grid_file)) {
    message("Prediction grid for ", survey_region, " (", sp_name, ") already exists.")
    return(NULL)
  }
  
  # Create a prediction grid for the specific survey region and species name
  ts.sf <- ts %>%
    dplyr::select(X, Y, eq_latitude, eq_longitude, crs) %>%
    mutate(X = X * 1000, Y = Y * 1000) %>% 
    st_as_sf(coords = c("X", "Y"), crs = unique(ts$crs))
  
  convex_hull <- concaveman(ts.sf, concavity = 3)
  convex_buff <- st_buffer(convex_hull, dist = 15e3)
  
  grid_geom <- st_make_grid(convex_buff, cellsize = 10e3, what = "centers") %>%
    st_intersection(convex_buff)
  
  bs.grid <- do.call(rbind, st_geometry(grid_geom)) %>%
    as_tibble() %>% 
    setNames(c("X", "Y")) %>%
    mutate(X = X / 1e3, Y = Y / 1e3)
  
  # Use only the years present in the original df
  bs.grid.yrs <- replicate_df(bs.grid, "sample_year", unique(ts$sample_year))
  # if the formula uses doy2 then use this one, if not use scale doy (meidan scaled)
  has_doy2 <- any(str_detect(as.character(fit$formula[[1]]), "scale_doy2"))
  
  if (has_doy2) {
    bs.grid.doy <- replicate_df(bs.grid.yrs, "scale_doy2", ts_scaled_median)
  } else {
    bs.grid.doy <- replicate_df(bs.grid.yrs, "scale_doy", 0)
  }
  
  write.csv(bs.grid.doy, file = grid_file, row.names = FALSE)
  message("Created prediction grid here:", grid_file)
}

## wrapper function for sdmTMB

set_fit_TMB <- function(
    data,
    survey_region,
    sp_name,
    run_fast = FALSE,
    n_x = 200,
    ...) {
  
  set.seed(2026)
  today <- Sys.Date()
  year <- year(today)
  
  # set directory
  path <- here("Results", paste0(year, "_ESR"), paste0(sp_name, "_", survey_region), today)
  if(!dir.exists(path)) dir.create(path, recursive = T)
  
  # change knots and if running quickly
  # (i.e. not a final model), honestly sdmtMB is so fast i dont use this often
   if (run_fast) {
    n_x <- 50
  } else if (!survey_region %in% c("NBS", "SEBS")) {
    stop("Invalid survey region specified.")
  }
  
  
  # change data pending region & sp name
  if (!sp_name %in% names(data)) stop("Species not found in data list")
  
  ts <- data[[sp_name]] %>%
    filter(region == survey_region) %>%
    as.data.frame()
  
  if (nrow(ts) == 0) stop("No data supplied for region: ", survey_region)

  
  # # print out median doy for *knowledge*
  ts_median <- unique(ts$median_doy)
  ts_scaled_median <- unique(ts$scaled_median)
  message(survey_region, " ", sp_name, " median doy: ", ts_median)
  
  ## create mesh
  mesh <- make_mesh(ts, xy_cols = c("X", "Y"), n_knots = n_x)

  ## fit model 
  fit <- sdmTMB(data = ts, mesh = mesh, ...)
  
  # pulled from lauren's code
  # is the model fit?
  if(inherits(fit, "try-error")) {
    print(paste("model error for species", sp_name, "in region", survey_region))
    return(fit)
  }
  
  # if sanity fails then run optimization 
  if(!sanity(fit)$all_ok) {
    fit <- try(run_extra_optimization(fit, nlminb_loops = 2, newton_loops = 1))
  }
  
  # see if its better
  if(inherits(fit, "try-error")) {
    print(paste("model error for species", sp_name, "in region", survey_region)) 
    return(fit)
  }
  
  # save model object
  fam_label <- paste(fit$family$family, collapse = "_")
  saveRDS(fit, file = paste0(here(path, sp_name), fam_label, "_fit.RDS"))
  
  # create prediction grid
  create_prediction_grid(ts, fit, survey_region, sp_name, year, ts_scaled_median)
  
  message("Model fits saved to:", path)
  
  return(fit)
  
}
```


## fit sdmTMB models

- note i don't run this chunk unless i'm sure i want to wait around forever
below there is an option to read in most recent model run
```{r eval=FALSE, include=FALSE}
# in comparison to VAST 
# ObsModel 2,3 -> family = delta_gamma(link1 = "logit", link2 = "log")?
# FieldConfig, first two elements are the spatial/spatiotemporal structure for the presence / absence part of the model. The second two elements are the spatial/spatiotemporal structure for the encounter model. 

# other considerations, share_range in sdmTMB
# scale / range parameters are independently estimated for each random field but sdmTMB shares it (with share_range = T)
TMB::openmp(n = 8) # can use parallel::detectCores to see how many cores available.
nbs.forage.fit <- set_fit_TMB(
                           # wrapper arguments
                           data = ts.list, 
                           survey_region = "NBS", 
                           sp_name = "Forage", 
                           run_fast = FALSE,
                           n_x = 200,
                           
                           # sdmtmb arguments
                           formula = catch_kg ~ 0 + scale_doy2 + as.factor(sample_year),
                           offset = "log_effort",
                           time = "sample_year",
                           family = tweedie(link = "log"),
                           spatial = "on",
                           spatiotemporal = "iid",
                           share_range = TRUE,
                           anisotropy = TRUE
                           )

nbs.pollock.fit <- set_fit_TMB(
                          # wrapper arguments
                           data = ts.list, 
                           survey_region = "NBS", 
                           sp_name = "Pollock_A0", 
                           run_fast = FALSE,
                           n_x = 200,
                           
                           formula = catch_kg ~ 0 + scale_doy2 + as.factor(sample_year),
                           time = "sample_year",
                           offset = "log_effort",
                           family = delta_gamma(link1 = "logit", link2 = "log"),
                           spatial = list("on", "on"),
                           spatiotemporal = list("iid", "iid"),
                           share_range = TRUE,
                           anisotropy = TRUE
                           )

nbs.herring.fit <- set_fit_TMB(data = ts.list,
                                survey_region = "NBS",
                                sp_name = "Pacific Herring_All",
                                run_fast = FALSE,
                                n_x = 200,
                               
                                # sdmtmb arguments
                                formula = catch_kg ~ 0 + scale_doy2 + as.factor(sample_year),
                                time = "sample_year",
                                offset = "log_effort",
                                family = tweedie(link = "log"),
                                spatial = "on",
                                spatiotemporal = "iid",
                                share_range = TRUE,
                                anisotropy = TRUE
                                ) 

nbs.capelin.fit <- set_fit_TMB(data = ts.list, 
                           survey_region = "NBS", 
                           sp_name = "Capelin_All", 
                           run_fast = FALSE,
                           n_x = 200,
                           
                           formula = catch_kg ~ 0 + scale_doy2 + as.factor(sample_year),
                           time = "sample_year",
                           offset = "log_effort",
                           family = tweedie(link = "log"),
                           spatial = "on",
                           spatiotemporal = "iid",
                           share_range = TRUE,
                           anisotropy = TRUE
                           )

nbs.jellyfish.fit <- set_fit_TMB(data = ts.list,
                                 survey_region = "NBS",
                                 sp_name = "Jellyfish",
                                 run_fast = FALSE,
                                 n_x = 200,
                                 
                                 formula = catch_kg ~ 0 + scale_doy2 + as.factor(sample_year),
                                 time = "sample_year",
                                 offset = "log_effort",
                                 family = tweedie(link = "log"),
                                 spatial = "on",
                                 spatiotemporal = "iid",
                                 share_range = TRUE,
                                 anisotropy = TRUE
                                 )

nbs.pcod.fit <- set_fit_TMB(data = ts.list,
                                 survey_region = "NBS",
                                 sp_name = "Pacific Cod_A0",
                                 run_fast = FALSE,
                                 n_x = 100,
                                 
                                 formula = catch_kg ~ 0 + scale_doy2 + as.factor(sample_year),
                                 time = "sample_year",
                                 offset = "log_effort",
                                 family = tweedie(link = "log"),
                                 spatial = "off",
                                 spatiotemporal = "iid",
                                 share_range = TRUE,
                                 anisotropy = TRUE
                                 )

sebs.forage.fit <- set_fit_TMB(data = ts.list, 
                           survey_region = "SEBS", 
                           sp_name = "Forage", 
                           run_fast = FALSE,
                           n_x = 100,
                           
                           formula = catch_kg ~ 0 + scale_doy2 + as.factor(sample_year),
                           time = "sample_year",
                           offset = "log_effort",
                           family = tweedie(link = "log"),
                           spatial = "on",
                           spatiotemporal = "iid",
                           share_range = TRUE,
                           anisotropy = TRUE
                           )

sebs.pollock.fit <- set_fit_TMB(data = ts.list, 
                           survey_region = "SEBS", 
                           sp_name = "Pollock_A0", 
                           run_fast = FALSE,
                           n_x = 100,
                           
                           formula = catch_kg ~ 0 + scale_doy2 + as.factor(sample_year),
                           time = "sample_year",
                           offset = "log_effort",
                           family = tweedie(link = "log"),
                           spatial = "on",
                           spatiotemporal = "iid",
                           share_range = TRUE,
                           anisotropy = TRUE
                           )

sebs.herring.fit <- set_fit_TMB(data = ts.list, 
                           survey_region = "SEBS", 
                           sp_name = "Pacific Herring_All", 
                           run_fast = FALSE,
                           n_x = 100,
                           
                           formula = catch_kg ~ 0 + scale_doy2 + as.factor(sample_year),
                           time = "sample_year",
                           offset = "log_effort",
                           family = tweedie(link = "log"),
                           spatial = "on",
                           spatiotemporal = "iid",
                           share_range = TRUE,
                           anisotropy = TRUE
                           )

## fit model without 2016 
sebs.capelin.fit.red <- set_fit_TMB(data = ts.list, 
                           survey_region = "SEBS", 
                           sp_name = "Capelin_All", 
                           run_fast = FALSE,
                           n_x = 100,
                           
                           formula = catch_kg ~ 0 + scale_doy2 + as.factor(sample_year),
                           time = "sample_year",
                           offset = "log_effort",
                           spatial = list("on","off"),
                           spatiotemporal = list("iid", "iid"),
                           family = delta_gamma(link1 = "logit", link2 = "log"),
                           share_range = TRUE,  
                           anisotropy = TRUE
                                )

sebs.jellyfish.fit <- set_fit_TMB(data = ts.list,
                                 survey_region = "SEBS",
                                 sp_name = "Jellyfish",
                                 run_fast = FALSE,
                                 n_x = 200,
                                 
                                 formula = catch_kg ~ 0 + scale_doy2 + as.factor(sample_year),
                                 time = "sample_year",
                                 offset = "log_effort",
                                 family = tweedie(link = "log"),
                                 spatial = "on",
                                 spatiotemporal = "iid",
                                 share_range = TRUE,
                                 anisotropy = TRUE
                                 )

sebs.pcod.fit <- set_fit_TMB(data = ts.list,
                                 survey_region = "SEBS",
                                 sp_name = "Pacific Cod_A0",
                                 run_fast = FALSE,
                                 n_x = 200,
                                 
                                 formula = catch_kg ~ 0 + scale_doy2 + as.factor(sample_year),
                                 time = "sample_year",
                                 offset = "log_effort",
                                 family = tweedie(link = "log"),
                                 spatial = "on",
                                 spatiotemporal = "iid",
                                 share_range = TRUE,
                                 anisotropy = TRUE
                                 )

TMB::openmp(n = 1)  # goes back to only one cores
```

## load other models (optional)
read in models if you've already run them but want to evaluate or look at the model objects

```{r eval=FALSE, include=FALSE}
run_date <- "2026-07-13"

# need to add in family of final model you want to pull in

# nbs.forage.fit <- readRDS(here("Results", paste0(year(Sys.Date()), "_ESR"), "Forage_NBS",
#                                run_date, "Foragetweedie_fit.RDS"))
# nbs.pollock.fit <- readRDS(here("Results", paste0(year(Sys.Date()), "_ESR"), "Pollock_A0_NBS",
#                                run_date, "Pollock_A0binomial_Gamma_fit.RDS"))
# nbs.herring.fit <- readRDS(here("Results", paste0(year(Sys.Date()), "_ESR"), "Pacific Herring_All_NBS",
#                                run_date, "Pacific Herring_Alltweedie_fit.RDS"))
# nbs.capelin.fit <- readRDS(here("Results", paste0(year(Sys.Date()), "_ESR"), "Capelin_All_NBS",
#                                run_date, "Capelin_Alltweedie_fit.RDS"))
# nbs.jellyfish.fit <- readRDS(here("Results", paste0(year(Sys.Date()), "_ESR"), "Jellyfish_NBS",
#                                run_date, "Jellyfishtweedie_fit.RDS"))
# nbs.pcod.fit <- readRDS(here("Results", paste0(year(Sys.Date()), "_ESR"), "Pacific Cod_A0_NBS",
#                                run_date, "Pacific Cod_A0tweedie_fit.RDS"))
# 
# sebs.forage.fit <- readRDS(here("Results", paste0(year(Sys.Date()), "_ESR"), "Forage_SEBS",
#                                run_date, "Foragetweedie_fit.RDS"))
# sebs.pollock.fit <- readRDS(here("Results", paste0(year(Sys.Date()), "_ESR"), "Pollock_A0_SEBS",
#                                run_date, "Pollock_A0tweedie_fit.RDS"))
# sebs.herring.fit <- readRDS(here("Results", paste0(year(Sys.Date()), "_ESR"), "Pacific Herring_All_SEBS",
#                                run_date, "Pacific Herring_Alltweedie_fit.RDS"))
# sebs.capelin.fit.red <- readRDS(here("Results", paste0(year(Sys.Date()), "_ESR"), "Capelin_All_SEBS",
#                                run_date, "Capelin_Allbinomial_Gamma_fit.RDS"))
# sebs.jellyfish.fit <- readRDS(here("Results", paste0(year(Sys.Date()), "_ESR"), "Jellyfish_SEBS",
#                                run_date, "Jellyfishtweedie_fit.RDS"))
# sebs.pcod.fit <- readRDS(here("Results", paste0(year(Sys.Date()), "_ESR"), "Pacific Cod_A0_SEBS",
#                                run_date, "Pacific Cod_A0tweedie_fit.RDS"))

```


## model evaluation
From the model fitting object the sanity check has already been done, but here
we'll review the diffferent spatial and randomized quantiles
```{r}
# model list:
model.list <- list(nbs.forage.fit, nbs.pollock.fit, nbs.herring.fit, nbs.capelin.fit,
                   nbs.jellyfish.fit, nbs.pcod.fit,
                   sebs.forage.fit, sebs.pollock.fit, sebs.herring.fit, sebs.capelin.fit.red,
                   sebs.jellyfish.fit, sebs.pcod.fit)
names(model.list) <- c("nbs.forage", "nbs.pollock", "nbs.herring", "nbs.capelin",
                       "nbs.jellyfish", "nbs.pcod",
                       "sebs.forage", "sebs.pollock", "sebs.herring", "sebs.capelin",
                       "sebs.jellyfish", "sebs.pcod")
```


this takes longer to run since i'm simulating 500 times to then extract dharma residuals.
```{r}
# include run_date if you are loading models that were run a different day than today
# run_date <- "2026-07-13"
# create a list of data sep by sp and region based on the original input to the fit sdmtMBs
tmp <- dplyr::bind_rows(ts.list)
ts.list.region <- split(tmp, list(tmp$sp_name, tmp$region))

names(ts.list.region) <- c("nbs.capelin","nbs.forage", "nbs.jellyfish", 
                           "nbs.pcod", "nbs.herring", "nbs.pollock",
                           "sebs.capelin", "sebs.forage", "sebs.jellyfish", 
                           "sebs.pcod", "sebs.herring", "sebs.pollock")

# testing value
#name <- "nbs.capelin.fit"
# create empty plot list
residual.plots <- list()
spatial.plots <- list()
residual.df <- list()


for(name in names(model.list)) {
  mod <- model.list[[name]]
  df <- ts.list.region[[name]]
  
  # get model names
  mod_fam <- mod$family$family
  fam_label <- paste(mod_fam, collapse = "_")
  
  # extract residuals
  sims <- simulate(mod, nsim = 500, type = "mle-mvn")
  resid <- dharma_residuals(sims, mod, return_DHARMa = TRUE)
  
  # create dharma plots as ggplots
  p_qq <- as.grob(~DHARMa::plotQQunif(resid))
  p_res <- as.grob(~DHARMa::plotResiduals(resid))
  
  # combine residual plots
  dharma_plots <- wrap_plots(p_qq, p_res) +
    plot_annotation(title = paste(name, fam_label, "DHARMa Diagnostics"))
  
  # put plot in a list
  residual.plots[[name]] <- dharma_plots
  
  # save location directory creation ifelse
  save_path <- here("Results",
                    paste0(year(Sys.Date()), "_ESR"),
                    paste0(unique(df$sp_name), "_", unique(df$region)), run_date)
  if(!dir.exists(save_path)) dir.create(save_path, recursive = TRUE)
  
  # save residual plot
  ggsave(filename = file.path(save_path, paste0(unique(df$sp_name), "_", fam_label, "_resid_plot.png")),
         plot = dharma_plots, width = 10, height = 5)
  
  # put the residuals in a df for spatial plotting
  df$resid <- resid$scaledResiduals
  residual.df[[name]] <- df
  
  # create spatial plot
  spatials <- ggplot(df) +
    geom_point(aes(X,Y, col = resid)) +
    labs(title = paste(name, fam_label, "Spatial residuals")) +
    scale_colour_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0.5) +
    facet_wrap(~sample_year) +
    coord_fixed() +
    theme_minimal()
  
  # save plot
  ggsave(filename = here("Results", paste0(year(Sys.Date()), "_ESR"), 
                         paste0(unique(df$sp_name), "_", unique(df$region)), run_date,
                         paste0(unique(df$sp_name), fam_label, "_spatial_resid.png")),
         plot = spatials, width = 8, height = 8)
  
  # put into list to look at
  spatial.plots[[name]] <- spatials
  


  print(name)
  residual.df[[name]] <- df
  
}
```

#### review residual plots
review the residual plots and spatial plots for each fitted object
```{r}
residual.plots[[1]] 
sanity(nbs.forage.fit)
print(tidy(nbs.forage.fit, effects = "fixed"), n = 22) # pretty good
tidy(nbs.forage.fit, effects = "ran_pars")
residual.plots[[2]]
sanity(nbs.pollock.fit)
print(tidy(nbs.pollock.fit, effects = "fixed"), n = 22) # std.errors look okay
tidy(nbs.pollock.fit, effects = "ran_pars")
residual.plots[[3]] # this looks good
residual.plots[[4]] # this isn't my fav - switching to delta-lognormal with iid & spatial on x2
residual.plots[[5]] # not so nice variation fro the tweedie distribution, should we be using different residuals? 
residual.plots[[6]] # same
residual.plots[[7]] # looks good for a hurdle model
residual.plots[[8]] # not bad not great


# spatial residauls, looking for clustering of similar direction poitns,
# used scaled residuals from the dharma object
spatial.plots[[1]]
spatial.plots[[2]] # 
spatial.plots[[3]]
spatial.plots[[4]] # nbs capelin very few high abundance caught
spatial.plots[[5]]
spatial.plots[[6]] # sebs pollock, maybe some negative residuals in the northern part of the survey early on? Part of that is the disappearance of the cold pool? 
spatial.plots[[7]]
spatial.plots[[8]]
```



# Predict biomass across prediction grid and calculate index biomass
Read in correct grid/fitted models
```{r eval=FALSE, include=FALSE}
# here is where you'd add another species if one was added to this script
model.opts <- data.frame(sp_name = c("Forage","Pollock_A0", "Pacific Herring_All",
                                  "Capelin_All", "Jellyfish", "Pacific Cod_A0"),
                      model_names = c("forage", "pollock", "herring", "capelin",
                                      "jellyfish", "pcod"))

# maybe best to remove the nbs pcod model
model.list[["nbs.pcod"]] <- NULL

#predict & extract indicies
predict_index <- function(model.name, model.list, model_options = model.opts) {
  mod <- model.list[[model.name]]
  
  # determine region from model.name
  if (str_detect(model.name, "(?i)^nbs")) {
    region <- "NBS"
  } else if (str_detect(model.name, "(?i)^sebs")) {
    region <- "SEBS"
  } else {
    stop("Could not determine region from model name: ", model.name)
  }
  
  # match the lowercase model name with a single search list anywhere in the model.name
  shorthand_sp <- str_extract(tolower(model.name), paste(model_options$model_names, collapse = "|"))
  # if its an na it explodes
  if (is.na(shorthand_sp)) {
    stop("No matching shorthand sp in:", model.name)
  }
  
  # filter to correct sp_name
  full_sp_name <- model_options %>%
    filter(model_names == shorthand_sp) %>%
    pull(sp_name)
  
  # load correct prediction grid
  grid_path <- here("Results", 
                    paste0(year(Sys.Date()), "_ESR"),
                    "Prediction_grids",
                    paste0(region, "_", full_sp_name, "_prediction-grid.csv"))
  # check to make sure it exists
  if(!file.exists(grid_path)){
    stop("grid does not exist", grid_path)
  }
  
  grid <- read.csv(grid_path)
  
  # translate the species options
  # predictions and index take some time to run
  preds <- predict(mod, newdata = grid, return_tmb_object = TRUE)
  index <- get_index(preds, bias_correct = TRUE, area = 1)
  
  # column with model name
  index$model <- model.name
  print(model.name)
  return(index)
  
}

# if you want to use parallel processing with increased cores
#TMB::openmp(n = 8) # can use parallel::detectCores to see how many cores the computer has

index_results <- map(names(model.list), 
    ~ predict_index(model.name = .x, model.list = model.list))

indicies <- bind_rows(index_results) %>%
  mutate(region = ifelse(str_detect(model, "^nbs\\."), "NBS", "SEBS"),
         species = ifelse(str_detect(model, "forage"), "Forage", 
                          ifelse(str_detect(model, "pollock"), "Pollock",
                                           ifelse(str_detect(model, "capelin"), "Capelin",
                                                             ifelse(str_detect(model, "herring"), "Herring",
                                                                    ifelse(str_detect(model, "pcod"), "Pacific Cod",
                                                                           ifelse(str_detect(model, "jellyfish"), "Jellyfish", NA)))))))

# once we determine which capelin model is correct add it into the indicies df
# write.csv(indicies, here("Results",
#                          paste0(year(Sys.Date()), "_ESR"),
#                          paste0("Index_rel_biomass_", Sys.Date(), ".csv")), row.names = F)
# TMB::openmp(n = 1) 
```

# plot the ESR indicies

```{r}
# if they need to be loaded can be done here
# include the correct date of the version of relative abundance  you want to use
correct.date <- "2026-07-13" # most recent run
indicies <- read.csv(here("Results", paste0(year(Sys.Date()), "_ESR"),
              paste0("Index_rel_biomass_", correct.date, ".csv"))) %>%
  group_by(region, species) %>%
  complete(sample_year = 2003:year(Sys.Date())) %>%
  ungroup() %>%
  filter(model != "nbs.pcod") # remove this becasue it doesn't converge

sp.list <- unique(indicies$species)

plot_ESR <- function(sp_name, data) {
  dat <- filter(data, species == sp_name)
  
  ggplot(data = dat, aes(x = sample_year, y = est/1000, group = region, 
                          color = region)) +
          geom_pointrange(aes(ymin = lwr/1000, ymax = upr/1000, 
                              group = region, color = region))+
          geom_line() +
          #geom_errorbar(aes(ymin = lwr/1000, ymax = upr/1000,
           #                 group = region, color = region), width=.1, 
            #            position=position_dodge(0.05)) +
          theme_cowplot(12)+
          ylab(bquote("Relative Biomass Index (metric tonne/"* ~ km^2*')'))+
          labs(title=paste(sp_name, "- Late Summer/Fall Surface Trawl Surveys"), 
               x ="Year", color = "Region", cex=3)+
          theme(legend.position = "right",
                axis.title = element_text(size = 16),
                legend.text = element_text(size = 12),
                legend.title = element_text(size = 16),
                axis.text = element_text(size = 12),
                panel.background = element_rect(colour = "black", linewidth=1))+
          scale_x_continuous(breaks = seq(min(dat$sample_year), max(dat$sample_year), 5)) +
          #scale_y_continuous(limits = c(0, 650)) +
          scale_color_manual(values=c('#003366','#0099CC'))#+
          #scale_color_manual(values=c('#003366','#0099CC'))+
          #geom_point(size=1)
  
}


plot_list <- list()
for(i in 1:length(sp.list)) {
  sp_name <- sp.list[[i]]
  plot_list[[i]] <- plot_ESR(sp_name = sp_name, data = indicies)
}

gridExtra::grid.arrange(grobs = plot_list)

# check to make sure this naming is correct before saving
names(plot_list) <- c("Capelin_All", "Forage", "Pacific Herring_All", 
                      "Jellyfish", "Pollock_A0", "Pacific Cod_A0")

# save the esr plots
lapply(1:length(plot_list), 
       function(x)ggsave(filename = here("Results", paste0(year(Sys.Date()), "_ESR"),
                                         paste0(names(plot_list)[[x]], 
                                                fam_label, "_timeseries_", run_date, ".pdf")),
                           dpi = 300, height = 5, width = 6.5,
                         plot = plot_list[[x]]))


```


Still to do - use the ESR facet plot template to calculate timeseries mean and standard deviation to show index variation around mean. 
Template code below. Needs to be adjusted to ours.
Plot NBS and SEBS separately. 

```{r}
#Code to create standardized multi-facet time series plots for ESRs
#Code largely based on GAP Products figure code (https://github.com/afsc-gap-products/esrindex) POC: Sean Rohan


library(ggplot2)
library(dplyr)
library(data.table)

plank=read.csv("plank.csv")

##change wide format to long format table
plank.long <- melt(setDT(plank), id.vars = c("Year"), variable.name = "group")

## calculate time series mean and sd for each species
TS_sum <- plank.long %>%
  group_by(group) %>%
  summarise(mean = mean(value, na.rm=T),sd = sd(value, na.rm=T))

## blue-strip color theme from gap-products/esrindex
theme_blue_strip <- function() {
  theme_bw() %+replace%
    theme(axis.title = element_text(color = "black", face = "bold"),
          axis.text = element_text(color = "black"),
          axis.ticks = element_line(color = "black"),
          panel.grid = element_blank(),
          legend.title = element_blank(),
          legend.position = "bottom",
          strip.text = element_text(size = 9,
                                    color = "white",
                                    face = "bold",
                                    margin = margin(0.5, 0, 0.5, 0, "mm")),
          strip.background = element_rect(fill = "#0055a4",
                                          color = NA))
}

## define other variables used in gap-products/esrindex code
n_facet_row <- length(unique(plank.long$group))
lab_fun <- scales::label_comma
lab_trim <- FALSE
lab_digits <- NULL

#customize labels for each subplot 

# plot labels
new_labels.f=c("plank1.Mean"="plankton sp.1","plank2.Mean"="plankton sp.2","plank3.Mean"="plankton sp.3", "plank.Total"="Total Plankton")

##############
##plot 

tiff("Fig_Plank_2025.tiff", width=7, height=7, units="in", res=600) 

#gapproducts plot -note commented out code for CI ribbons and errorbars has old GAP-products naming conventions;  

ggplot() +
  #create horizontal lines of long-term mean and +/- 1SD
  geom_hline(data = TS_sum,
             mapping = aes(yintercept = mean),
             color = "black") +
  geom_hline(data = TS_sum,
             mapping = aes(yintercept = mean+sd), 
             linetype = 2,
             color = "black") +
  geom_hline(data = TS_sum,
             mapping = aes(yintercept = mean-sd), 
             linetype = 2,
             color = "black") +

  #plot confidence level ribbons or error bars (update data name and variable names pred_lci,obs_lci ) 
  #geom_ribbon(data = fit_dat,
  #            mapping = aes(x = year,
  #                          ymin = pred_lci,
  #                          ymax = pred_uci),
  #            alpha = 0.3,
  #            fill = ribbon_fill) +
  #geom_errorbar(data = fit_dat,
  #              mapping = aes(x = year,
  #                            ymin = obs_lci,
  #                            ymax = obs_uci), 
  #              width = 0.5,
#              color = errorbar_color) +
  
  #plot data points
geom_line(data = plank.long,mapping = aes(x = Year, y = value),
          color = "black") +
  geom_point(data = plank.long,
             mapping = aes(x = Year, y = value),
             color = "black") +
  #axis labels and scales
  ylab(~paste("Plankton mean biomass (g ∙ ",m^-3,")"))+
  scale_y_continuous( expand = expansion(mult = c(0, 0.5)), 
                      labels = lab_fun(trim = lab_trim)) + #if want extra decimal point mult = c(0, 0.05)
  scale_x_continuous(name = "Year")+#,
  expand_limits(y = 0) +
  facet_wrap(~group, 
             scales = "free",
             nrow = n_facet_row, labeller = labeller(group = new_labels.f)) + # scales = "free_y"
  theme_blue_strip()

dev.off() 

```

