---
title: "fit_sdm_models"
output: html_document
date: "2026-06-25"
editor_options: 
  chunk_output_type: console
---

Pull in fxns
```{r}
source("Scripts/helper_fxns/clean_pull_data.R")
```

# libraries
```{r}
# make sure the proper packages are install and loaded
pkgs <- c("visreg", "sdmTMB", "sf", "concaveman", "cowplot", "gridExtra", "patchwork",
          "lubridate", "here")

for (pkg in pkgs) {
  if (!require(pkg, character.only = TRUE)) {
    install.packages(pkg)
    library(pkg, character.only = TRUE)
  } else {
    message(paste("Package '", pkg, "' is already installed and loaded.", sep = ""))
  }
}
```

# Pull in data
```{r}
tsns <- c(934083, 161979, 162035, 161980, 161976, 161977, 551209, 161975, 164711, 162041, 164708, 171672, 171671) # right now as i have this set up the tsns have to be these. Otherwise i need to adjust how they are combined in the clean_pull_data function
this.year <- 2025

nbs.ts <- pull_clean_catch(this.year = this.year, tsns = tsns, region = "NBS") %>%
  # this pivot longer wont work if you change the tsns above 
  pivot_longer(cols = c("Chinook Salmon_J":"Forage"), names_to = "sp_name", values_to = "catch_kg") %>%
  add_utm_columns(., ll_names = c("Lon", "Lat"),
                  utm_names = c("X", "Y"), ll_crs = 4326, units = "km") %>%
  mutate(region = "NBS",
         crs = 32603)

sebs.ts <- pull_clean_catch(this.year = this.year, tsns = tsns, region = "SEBS") %>%
  # this pivot longer wont work if you change the tsns above 
  pivot_longer(cols = c("Chinook Salmon_J":"Forage"), names_to = "sp_name", values_to = "catch_kg") %>%
  add_utm_columns(., ll_names = c("Lon", "Lat"),
                  utm_names = c("X", "Y"), ll_crs = 4326, units = "km") %>%
  mutate(region = "SEBS", 
         crs = 32602)

# glimpse(nbs.ts)
# glimpse(sebs.ts)
# combine and then split
ts <- rbind(sebs.ts, nbs.ts) # crs for NBS is 32603, SEBS is 32602
  
# get the species names in a vector, there are 31, but we're only interested in a few for the ESRs 
species_names <- unique(ts$sp_name) 
species_names <- species_names[species_names %in% c("Capelin_All", "Forage", "Pacific Herring_All",
                                                   "Pollock_A0")]
ts <- filter(ts, sp_name %in% species_names) %>%
  filter(!(is.na(effort))) # drop any stations that have NA for effort

ts <- mutate(ts, log_effort = log(effort)) %>%
  group_by(region, sp_name) %>%
  mutate(median_doy = median(doy),
         scale_doy = doy - median_doy,
         mean_doy = mean(doy),
         sd_doy = sd(doy),
         scale_doy2 = (doy - mean_doy)/sd_doy,
         scaled_median = (median_doy - mean_doy) / sd_doy)

ts.red <- filter(ts, !(region == "SEBS" & sp_name == "Capelin_All" & sample_year == 2016))
#write.csv(ts.red, here("Results", paste0(year(Sys.Date()), "_ESR"), paste0("nbs_sebs_timeseries"))

# datalist just by the species name so includes region for both within a specis
ts.list <- split(ts.red, list(ts.red$sp_name))

```

## understand the catch / encounter data for each region and species
Things to look for -
- 0% or 100% encounter rate for years 

```{r}
# check to make sure there's no negative or NAs in catch data
table(is.na(ts$catch_kg)) # none
table(ts$catch_kg < 0) # none
table(is.na(ts$effort)) # any NAs in effort need to be removed

forage <- ts.list[["Forage"]]
pol <- ts.list[["Pollock_A0"]]
capelin <- ts.list[["Capelin_All"]]
herring <- ts.list[["Pacific Herring_All"]]

# both regions here
# some years with zero, 0 encounter rates, i.e., multiple years where all stations had some forage catch
# foro NBS very few zero encounters, SEBS more 0 encounters but very few
table(forage$catch_kg > 0, forage$sample_year, forage$region)
# for NBS good mix of pol encounter/non encounter, for SEBS years where no surveys done but otherwise a good mix in early years for encounter non encounter, but later years only positive encounter mainly (reduced survey footprint meaning only middle domain stations with pollock)
table(pol$catch_kg > 0, pol$sample_year, pol$region)
# for NBS good mix of capelin encounter/non encounter in early years, with later years a lot fewer capelin encountered (especially 2017-2021). For SEBS, even fewer encounters across the board especially in recent years (2016 no encounters etc)
table(capelin$catch_kg > 0, capelin$sample_year, capelin$region)
# for NBS good mix of encounter non encounter for both NBS and SEBS
table(herring$catch_kg >0, herring$sample_year, herring$region)


#' conclusions:
#' Forage: perhaps no need for a hurdle model, just use a positive encounter model for catch such as the tweedie
#' Pollock: NBS likely needs a hurdle model, like delta_gamma or delta_lognormal, SEBS can be tweedie
#' Capelin, high zero encounters probably delta_lognormal
#' Herring probably can use a delta-gamma or delta_lognormal. 

print(pol %>% filter(region == "NBS",
               catch_kg > 0) %>%
  count(sample_year), n = 21)
```


# sdmTMB wrapper fxn
```{r}
# this function fits a sdmTMB model that has offset log effort and a model structure of f-year and s-scale.doy
# this has indepedent year effects
set_fit_TMB <- function(
    data,
    survey_region,
    sp_name,
    run_fast = FALSE,
    n_x = 200,
    ...) {
  
  set.seed(2026)
  
  # set directory for results
  today <- Sys.Date()
  year <- year(today)
  path <- here("Results", paste0(year, "_ESR"), paste0(sp_name, "_", survey_region), today)
  if(!dir.exists(path)) dir.create(path, recursive = T)
  
  # change knots and if running quickly
  # (i.e. not a final model)
   if(run_fast == TRUE){
    n_x <- 50
  } else {
    # if we want to have different models
    if(survey_region == "NBS") {
      n_x = 200
    } else {
      if(survey_region == "SEBS") {
        n_x = 200
      }
    }
  }
  
  
  # change data pending region & sp name
  if(survey_region == "NBS"){
    ts.tmp <- data[[sp_name]]
    ts <- filter(ts.tmp, region == survey_region) %>%
      as.data.frame()
  } else {
    if(survey_region == "SEBS") {
      ts.tmp <- data[[sp_name]]
    ts <- filter(ts.tmp, region == survey_region) %>%
      as.data.frame()
    } else {stop("data not supplied for region")}
    
  }
  
  # # print out median doy for *knowledge*
  ts_median <- unique(ts$median_doy)
  ts_scaled_median <- unique(ts$scaled_median)
  
  print(paste(survey_region, sp_name, "median doy", ts_median))
  
  ## create mesh
  mesh <- make_mesh(ts, 
            xy_cols = c("X", "Y"),
            n_knots = n_x)

  ## fit model 
  fit <- sdmTMB(data = ts,
                mesh = mesh, ...)

  print(sanity(fit))
  
  mod_fam <- fit$family$family
  fam_label <- paste(mod_fam, collapse = "_")
  
  # pulled from lauren's code
  # is the model fit?
  if(inherits(fit, "try-error")) {
    print(paste("model error for species", sp_name, "in region", survey_region))
    
  } else {
    # otherwise, check to make sure the sanity checks are met
    if(sanity(fit)$all_ok == FALSE) fit <- try(run_extra_optimization(fit, nlminb_loops = 2, 
                                                                      newton_loops = 1)) }# close loop
  
  # after that, check to see if the
  if(inherits(fit, "try-error")) {
    print(paste("model error for species", sp_name, "in region", survey_region)) }
  else {
    # if the model doesn't have any sanity issues or model issues then... 
    
    # save model object
    saveRDS(fit, file = paste0(here(path, sp_name), fam_label, "_fit.RDS"))
    
    # look to see if there is a prediction grid already in this years ESR folder
    # Add a wrapper if statement if its the capelin prediction grid 
    if(sp_name == "Capelin_All" & survey_region == "SEBS") {
      # check to see if there's a capelin prediction grid
      if(file.exists(here("Results", paste0(year, "_ESR"),
                     paste0(survey_region, "_Capelin_All", "_prediction-grid.csv")))) {
         print(paste("prediction grid for", survey_region, sp_name, "already exists"))
      } else {
        # create sebs capelin prediction grid
        ts.sf <- ts %>%
        dplyr::select(c(X, Y, eq_latitude, eq_longitude)) %>%
        mutate(X = X*1000, Y = Y*1000) %>% # convert to meters for the crs conversion
        st_as_sf(coords = c("X", "Y"), crs = unique(ts$crs))
      
        convex_hull <- concaveman(ts.sf, concavity = 3)
        convex_buff <- st_buffer(convex_hull, dist = 15e3)
        grid <- st_make_grid(convex_buff, cellsize = 10e3, what = "centers") %>%
          st_intersection(convex_buff)
        
        # create spatial grid
        bs.grid <- do.call(rbind, st_geometry(grid)) %>%
          as.tibble() %>% setNames(c("X", "Y")) %>%
          # convert df m back to km
          mutate(X = X/1e3, Y = Y/1e3)
        # replicate the grid across years (always in model)
        bs.grid.yrs <- replicate_df(bs.grid, "sample_year", unique(ts$sample_year))
        # assuming we have scale_doy in model (which we should) replicate with the scaled doy
        # to predict on the median value of the survey
        
        # need to insert an if else in case you're using scale_doy2 versus median scaled
        # if there's any value matching scale_doy2 in the formula then we need to replicate_df
        # based on that value not scale_doy
        if(any(str_detect(as.character(fit$formula[[1]]), "scale_doy2")) == TRUE) {
          # if true use scale_doy2
          bs.grid.doy <- replicate_df(bs.grid.yrs, "scale_doy2", ts_scaled_median)
        } else {
          bs.grid.doy <- replicate_df(bs.grid.yrs, "scale_doy", 0)
        }
        
      
        write.csv(bs.grid.doy, file = here("Results", paste0(year, "_ESR"),
                       paste0(survey_region, "_Capelin_All", "_prediction-grid.csv")),
                  row.names = FALSE)

      } # close capelin prediction grid creation 
    } else { 
      
      # if the sp_name is not capelin then... 
      if(file.exists(here("Results", paste0(year, "_ESR"),
                                         paste0(survey_region, "_prediction-grid.csv")))) {
        print(paste("prediction grid for", survey_region, "already exists"))
      } else {
        # create prediction grids for this model
      # take the data input, convert to sf 
      ts.sf <- ts %>%
        dplyr::select(c(X, Y, eq_latitude, eq_longitude)) %>%
        mutate(X = X*1000, Y = Y*1000) %>% # convert to meters for the crs conversion
        st_as_sf(coords = c("X", "Y"), crs = unique(ts$crs))
      
      convex_hull <- concaveman(ts.sf, concavity = 3)
      convex_buff <- st_buffer(convex_hull, dist = 15e3)
      grid <- st_make_grid(convex_buff, cellsize = 10e3, what = "centers") %>%
        st_intersection(convex_buff)
      
      # create spatial grid
      bs.grid <- do.call(rbind, st_geometry(grid)) %>%
        as.tibble() %>% setNames(c("X", "Y")) %>%
        # convert df m back to km
        mutate(X = X/1e3, Y = Y/1e3)
      # replicate the grid across years (always in model)
      bs.grid.yrs <- replicate_df(bs.grid, "sample_year", unique(ts$sample_year))
      # assuming we have scale_doy in model (which we should) replicate with the scaled doy
      # to predict on the median value of the survey
      # need to insert an if else in case you're using scale_doy2 versus median scaled
        # if there's any value matching scale_doy2 in the formula then we need to replicate_df
        # based on that value not scale_doy
        if(any(str_detect(as.character(fit$formula[[1]]), "scale_doy2")) == TRUE) {
          # if true use scale_doy2
          bs.grid.doy <- replicate_df(bs.grid.yrs, "scale_doy2", ts_scaled_median)
        } else {
          bs.grid.doy <- replicate_df(bs.grid.yrs, "scale_doy", 0)
        }
      
      write.csv(bs.grid.doy, file = here("Results", paste0(year, "_ESR"),
                                         paste0(survey_region, "_prediction-grid.csv")),
                row.names = FALSE)
      } # close prediction grid if else creation
    } # close the else of if spname == capelin
      
  
    
    
    
  } # close checking model errors
  
  print(paste("Model object and prediction grid (by year and scaled_doy and space) saved to", paste0(here(path))))
  
  return(fit)
  
}


```

## fit sdmTMB models
```{r eval=FALSE, include=FALSE}
# ObsModel 2,3 -> family = delta_gamma(link1 = "logit", link2 = "log")?
# FieldConfig, first two elements are the spatial/spatiotemporal structure for the presence / absence part of the model. The second two elements are the spatial/spatiotemporal structure for the encounter model. 

# other considerations, share_range in sdmTMB
# scale / range parameters are independently estimated for each random field but sdmTMB shares it (with share_range = T)
TMB::openmp(n = 8) # can use parallel::detectCores to see how many cores available.
nbs.forage.fit <- set_fit_TMB(
                           # wrapper arguments
                           data = ts.list, 
                           survey_region = "NBS", 
                           sp_name = "Forage", 
                           run_fast = FALSE,
                           n_x = 200,
                           
                           # sdmtmb arguments
                           formula = catch_kg ~ 0 + scale_doy2 + as.factor(sample_year),
                           offset = "log_effort",
                           time = "sample_year",
                           family = tweedie(link = "log"),
                           spatial = "on",
                           spatiotemporal = "iid",
                           share_range = TRUE,
                           anisotropy = TRUE
                           )

nbs.pollock.fit <- set_fit_TMB(
                          # wrapper arguments
                           data = ts.list, 
                           survey_region = "NBS", 
                           sp_name = "Pollock_A0", 
                           run_fast = FALSE,
                           n_x = 200,
                           
                           formula = catch_kg ~ 0 + scale_doy2 + as.factor(sample_year),
                           time = "sample_year",
                           offset = "log_effort",
                           family = delta_gamma(link1 = "logit", link2 = "log"),
                           spatial = list("on", "on"),
                           spatiotemporal = list("iid", "iid"),
                           share_range = TRUE,
                           anisotropy = TRUE
                           )

nbs.herring.fit <- set_fit_TMB(data = ts.list,
                                survey_region = "NBS",
                                sp_name = "Pacific Herring_All",
                                run_fast = FALSE,
                                n_x = 200,
                               
                                # sdmtmb arguments
                                formula = catch_kg ~ 0 + scale_doy2 + as.factor(sample_year),
                                time = "sample_year",
                                offset = "log_effort",
                               family = tweedie(link = "log"),
                               spatial = "on",
                               spatiotemporal = "iid",
                                # family = delta_lognormal(link1 = "logit", link2 = "log"),
                                # spatial = list("on", "on"),
                                # #spatiotemporal = list("off", "off"), # meets sanity
                                # spatiotemporal = list("iid", "iid"),
                                share_range = TRUE,
                                anisotropy = TRUE
                                ) 

nbs.capelin.fit <- set_fit_TMB(data = ts.list, 
                           survey_region = "NBS", 
                           sp_name = "Capelin_All", 
                           run_fast = FALSE,
                           n_x = 200,
                           
                           formula = catch_kg ~ 0 + scale_doy2 + as.factor(sample_year),
                           time = "sample_year",
                           offset = "log_effort",
                           family = tweedie(link = "log"),
                           spatial = "on",
                           spatiotemporal = "iid",
                           share_range = TRUE,
                           anisotropy = TRUE
                           ) 

sebs.forage.fit <- set_fit_TMB(data = ts.list, 
                           survey_region = "SEBS", 
                           sp_name = "Forage", 
                           run_fast = FALSE,
                           n_x = 100,
                           
                           formula = catch_kg ~ 0 + scale_doy2 + as.factor(sample_year),
                           time = "sample_year",
                           offset = "log_effort",
                           family = tweedie(link = "log"),
                           spatial = "on",
                           spatiotemporal = "iid",
                           share_range = TRUE,
                           anisotropy = TRUE
                           )

sebs.pollock.fit <- set_fit_TMB(data = ts.list, 
                           survey_region = "SEBS", 
                           sp_name = "Pollock_A0", 
                           run_fast = FALSE,
                           n_x = 100,
                           
                           formula = catch_kg ~ 0 + scale_doy2 + as.factor(sample_year),
                           time = "sample_year",
                           offset = "log_effort",
                           family = tweedie(link = "log"),
                           spatial = "on",
                           spatiotemporal = "iid",
                           share_range = TRUE,
                           anisotropy = TRUE
                           )

sebs.herring.fit <- set_fit_TMB(data = ts.list, 
                           survey_region = "SEBS", 
                           sp_name = "Pacific Herring_All", 
                           run_fast = FALSE,
                           n_x = 100,
                           
                           formula = catch_kg ~ 0 + scale_doy2 + as.factor(sample_year),
                           time = "sample_year",
                           offset = "log_effort",
                           family = tweedie(link = "log"),
                           spatial = "on",
                           spatiotemporal = "iid",
                           share_range = TRUE,
                           anisotropy = TRUE
                           )

## fit model without 2016 
sebs.capelin.fit.red <- set_fit_TMB(data = ts.list, 
                           survey_region = "SEBS", 
                           sp_name = "Capelin_All", 
                           run_fast = FALSE,
                           n_x = 100,
                           
                           formula = catch_kg ~ 0 + scale_doy2 + as.factor(sample_year),
                           time = "sample_year",
                           offset = "log_effort",
                           spatial = list("on","off"),
                           spatiotemporal = list("iid", "iid"),
                           family = delta_gamma(link1 = "logit", link2 = "log"),
                           share_range = TRUE,  
                           anisotropy = TRUE
                                )

TMB::openmp(n = 1)  # goes back to only one cores
```

## load other models (optional)
read in models if you've already run them but want to evaluate or look at the model objects

```{r eval=FALSE, include=FALSE}
run_date <- "2026-07-09"

# need to add in family of final model you want to pull in

# nbs.forage.fit <- readRDS(here("Results", paste0(year(Sys.Date()), "_ESR"), "Forage_NBS",
#                                run_date, "Foragetweedie_fit.RDS"))
# nbs.pollock.fit <- readRDS(here("Results", paste0(year(Sys.Date()), "_ESR"), "Pollock_A0_NBS",
#                                run_date, "Pollock_A0binomial_Gamma.RDS"))
# nbs.herring.fit <- readRDS(here("Results", paste0(year(Sys.Date()), "_ESR"), "Pacific Herring_All_NBS",
#                                run_date, "Pacific Herring_Alltweedie_fit.RDS"))
# nbs.capelin.fit <- readRDS(here("Results", paste0(year(Sys.Date()), "_ESR"), "Capelin_All_NBS",
#                                run_date, "Capelin_Alltweedie_fit.RDS"))
# 
# sebs.forage.fit <- readRDS(here("Results", paste0(year(Sys.Date()), "_ESR"), "Forage_SEBS",
#                                run_date, "Foragetweedie_fit.RDS"))
# sebs.pollock.fit <- readRDS(here("Results", paste0(year(Sys.Date()), "_ESR"), "Pollock_A0_SEBS",
#                                run_date, "Pollock_A0tweedie_fit.RDS"))
# sebs.herring.fit <- readRDS(here("Results", paste0(year(Sys.Date()), "_ESR"), "Pacific Herring_All_SEBS",
#                                run_date, "Pacific Herring_Alltweedie_fit.RDS"))
# sebs.capelin.fit <- readRDS(here("Results", paste0(year(Sys.Date()), "_ESR"), "Capelin_All_SEBS",
#                                run_date, "Capelin_Allbinomial_Gamma_fit.RDS"))

```


## model evaluation
From the model fitting object the sanity check has already been done, but here
we'll review the diffferent spatial and randomized quantiles
```{r}
# model list: 
model.list <- list(nbs.forage.fit, nbs.pollock.fit, nbs.herring.fit, nbs.capelin.fit,
                   sebs.forage.fit, sebs.pollock.fit, sebs.herring.fit, sebs.capelin.fit.red)
names(model.list) <- c("nbs.forage", "nbs.pollock", "nbs.herring", "nbs.capelin",
                       "sebs.forage", "sebs.pollock", "sebs.herring", "sebs.capelin")
```


this takes longer to run since i'm simulating 500 times to then extract dharma residuals.
```{r}
capelin.only <- filter(ts, region == "SEBS" & sp_name == "Capelin_All") %>%
  filter(sample_year != 2016)
everything.else <- filter(ts, !(region == "SEBS" & sp_name == "Capelin_All"))

# create two seperate lists 
ts.list.region <- split(everything.else, list(everything.else$sp_name, everything.else$region))
ts.list.region$Capelin_All.SEBS <- capelin.only

names(ts.list.region) <- c("nbs.capelin","nbs.forage", "nbs.herring", "nbs.pollock",
                           "sebs.capelin", "sebs.forage", "sebs.herring", "sebs.pollock")
# testing value
#name <- "nbs.capelin.fit"
# create empty plot list
residual.plots <- list()
spatial.plots <- list()
residual.df <- list()


for(name in names(model.list)) {
  mod <- model.list[[name]]
  df <- ts.list.region[[name]]
  
  # get model names
  mod_fam <- mod$family$family
  fam_label <- paste(mod_fam, collapse = "_")
  
  # extract residuals
  sims <- simulate(mod, nsim = 500, type = "mle-mvn")
  resid <- dharma_residuals(sims, mod, return_DHARMa = TRUE)
  
  # set up saving the plot
  png(filename = here("Results", paste0(year(Sys.Date()), "_ESR"), 
                      paste0(unique(df$sp_name),"_", unique(df$region)),
                           run_date, paste0(unique(df$sp_name), fam_label, "_resid_plot.png")))
  plot(resid, title = paste(name, fam_label))
  dev.off()
  # record the plot to put in a list
  residual.plot <- recordPlot()
  plot.new()
  residual.plots[[name]] <- residual.plot
  
  
  
  # put the residuals in a df for spatial plotting
  df$resid <- resid$scaledResiduals
  
  
  # create spatial plot
  spatials <- ggplot(df) +
    geom_point(aes(X,Y, col = resid)) +
    labs(title = paste(name, fam_label)) +
    scale_colour_gradient2() +
    facet_wrap(~sample_year) +
    coord_fixed()
  
  # save plot
  ggsave(filename = here("Results", paste0(year(Sys.Date()), "_ESR"), paste0(unique(df$sp_name), "_",
                                                                               unique(df$region)),
                           run_date, paste0(unique(df$sp_name), fam_label, "_spatial_resid.png")),
         plot = spatials)
  
  # put into list to look at
  spatial.plots[[name]] <- spatials
  


  print(name)
  residual.df[[name]] <- df
  
}
```

#### review residual plots
review the residual plots and spatial plots for each fitted object
```{r}
residual.plots[[1]] 
sanity(nbs.forage.fit)
print(tidy(nbs.forage.fit, effects = "fixed"), n = 22) # pretty good
tidy(nbs.forage.fit, effects = "ran_pars")
residual.plots[[2]]
sanity(nbs.pollock.fit)
print(tidy(nbs.pollock.fit, effects = "fixed"), n = 22) # std.errors look okay
tidy(nbs.pollock.fit, effects = "ran_pars")
residual.plots[[3]] # this looks good
residual.plots[[4]] # this isn't my fav - switching to delta-lognormal with iid & spatial on x2
residual.plots[[5]] # not so nice variation fro the tweedie distribution, should we be using different residuals? 
residual.plots[[6]] # same
residual.plots[[7]] # looks good for a hurdle model
residual.plots[[8]] # not bad not great


# spatial residauls, looking for clustering of similar direction poitns,
# used scaled residuals from the dharma object
spatial.plots[[1]]
spatial.plots[[2]] # 
spatial.plots[[3]]
spatial.plots[[4]] # nbs capelin very few high abundance caught
spatial.plots[[5]]
spatial.plots[[6]] # sebs pollock, maybe some negative residuals in the northern part of the survey early on? Part of that is the disappearance of the cold pool? 
spatial.plots[[7]]
spatial.plots[[8]]
```



# Predict biomass across prediction grid and calculate index biomass
Read in correct grid/fitted models
```{r eval=FALSE, include=FALSE}
# read in prediction grid for said survey region 
nbs.grid <- read.csv(here("Results", 
                          paste0(year(Sys.Date()), "_ESR"), "NBS_prediction-grid.csv"))
sebs.grid <- read.csv(here("Results", 
                           paste0(year(Sys.Date()), "_ESR"), "SEBS_prediction-grid.csv"))

sebs.grid.red <- read.csv(here("Results", 
                               paste0(year(Sys.Date()), "_ESR"), "SEBS_Capelin_All_prediction-grid.csv"))

# predict sebs and nbs
# model list 
#head(model.list)
# remove the capelin from the model list
model.list[["sebs.capelin"]] <- NULL

grid.list <- list(nbs.grid, sebs.grid)
names(grid.list) <- c("nbs", "sebs")

#predict & extract indicies
predict_index <- function(model.name, model.list, grid.list) {
  mod <- model.list[[model.name]]
  
  #is there a nbs at beginning of the model name?
  if(str_detect(model.name, "^nbs\\.")) {
    # make the grid the nbs grid
    grid <- grid.list[["nbs"]]
  } else {
    grid <- grid.list[["sebs"]]
    
  }
  
  preds <- predict(mod, newdata = grid, return_tmb_object = TRUE)
  index <- get_index(preds, bias_correct = TRUE, area = 1)
  
  # column with model name
  index$model <- model.name
  print(model.name)
  return(index)
  
}

# if you want to use parallel processing with increased cores
#TMB::openmp(n = 8) # can use parallel::detectCores to see how many cores the computer has

index_results <- map(names(model.list), 
    ~ predict_index(model.name = .x, model.list = model.list, grid.list = grid.list))

indicies <- bind_rows(index_results) %>%
  mutate(region = ifelse(str_detect(model, "^nbs\\."), "NBS", "SEBS"),
         species = ifelse(str_detect(model, "forage"), "Forage", 
                          ifelse(str_detect(model, "pollock"), "Pollock",
                                           ifelse(str_detect(model, "capelin"), "Capelin",
                                                             ifelse(str_detect(model, "herring"), "Herring", NA)))))

# capelin model for sebs were fit differently than the other models
pred.capelin <- predict(sebs.capelin.fit.red, newdata = sebs.grid.red, return_tmb_object = T)

capelin.index <- get_index(pred.capelin, bias_correct = TRUE, area = 1) %>%
  mutate(region = "SEBS",
         species = "Capelin",
         model = "sebs.capelin")

indicies2 <- rbind(indicies, capelin.index)

# once we determine which capelin model is correct add it into the indicies df
# write.csv(indicies2, here("Results",
#                          paste0(year(Sys.Date()), "_ESR"),
#                          paste0("Index_rel_biomass_", Sys.Date(), ".csv")), row.names = F)
# TMB::openmp(n = 1) 
```

# plot the ESR indicies

```{r}
# if they need to be loaded can be done here
# include the correct date of the version of relative abundance  you want to use
correct.date <- "2026-07-10" # most recent run
indicies <- read.csv(here("Results", paste0(year(Sys.Date()), "_ESR"),
              paste0("Index_rel_biomass_", correct.date, ".csv"))) %>%
  group_by(region, species) %>%
  complete(sample_year = 2003:year(Sys.Date())) %>%
  ungroup()

sp.list <- unique(indicies$species)

plot_ESR <- function(sp_name, data) {
  dat <- filter(data, species == sp_name)
  
  ggplot(data = dat, aes(x = sample_year, y = est/1000, group = region, 
                          color = region)) +
          geom_pointrange(aes(ymin = lwr/1000, ymax = upr/1000, 
                              group = region, color = region))+
          geom_line() +
          #geom_errorbar(aes(ymin = lwr/1000, ymax = upr/1000,
           #                 group = region, color = region), width=.1, 
            #            position=position_dodge(0.05)) +
          theme_cowplot(12)+
          ylab(bquote("Relative Biomass Index (metric tonne/"* ~ km^2*')'))+
          labs(title=paste(sp_name, "- Late Summer/Fall Surface Trawl Surveys"), 
               x ="Year", color = "Region", cex=3)+
          theme(legend.position = "right",
                axis.title = element_text(size = 16),
                legend.text = element_text(size = 12),
                legend.title = element_text(size = 16),
                axis.text = element_text(size = 12),
                panel.background = element_rect(colour = "black", linewidth=1))+
          scale_x_continuous(breaks = seq(min(dat$sample_year), max(dat$sample_year), 5)) +
          #scale_y_continuous(limits = c(0, 650)) +
          scale_color_manual(values=c('#003366','#0099CC'))#+
          #scale_color_manual(values=c('#003366','#0099CC'))+
          #geom_point(size=1)
  
}


plot_list <- list()
for(i in 1:length(sp.list)) {
  sp_name <- sp.list[[i]]
  plot_list[[i]] <- plot_ESR(sp_name = sp_name, data = indicies)
}

gridExtra::grid.arrange(grobs = plot_list)

names(plot_list) <- c("Capelin_All", "Forage", "Pacific Herring_All", "Pollock_A0")

# save the esr plots
lapply(1:length(plot_list), 
       function(x)ggsave(filename = here("Results", paste0(year(Sys.Date()), "_ESR"),
                                         paste0(names(plot_list)[[x]], 
                                                fam_label, "_timeseries_", run_date, ".pdf")),
                           dpi = 300, height = 5, width = 6.5,
                         plot = plot_list[[x]]))


```



